### 2024-11-24 paper/paper.qmd
Rough draft of paper complete, with finalised clean data.

---
title: "Decoding UK Hospital Wait Times: Infrastructure, Workforce, and Demand"
subtitle: "How Systemic Pressures Shape Delays in Medical Care"
author: 
  - Chenika Bukes
thanks: "Code and data are available at: [https://github.com/chenikabukes/UnitedKingdomHealthcare](https://github.com/chenikabukes/UnitedKingdomHealthcare)."
date: today
date-format: long
abstract: "Hospital wait times in the United Kingdom are a critical measure of healthcare efficiency, reflecting the balance between resources and demand. This study examines the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for seven key medical procedures from 2015 to 2019. Using regression modeling, we find that reduced hospital beds per capita significantly prolong wait times,and while higher physician availability helps, it does little to offset the impact of declining bed availability. Additionally, Type 1 major care attendance rates correlate with increased delays, further indicating systemic bottlenecks. These findings highlight the urgent need for strategic resource allocation and demand management to address delays and improve patient outcomes."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(arrow)
library(gridExtra)
library(modelsummary)
```

# Introduction

Timely access to healthcare is fundamental to achieving optimal health outcomes. In the United Kingdom, the National Health Service (NHS) faces mounting pressure to manage increasing patient demands amidst resource constraints. Hospital wait times, a key performance indicator, have become a focal point for policymakers and stakeholders alike. Extended wait times not only compromise patient care but also signal underlying inefficiencies in healthcare delivery systems.

This paper investigates how hospital infrastructure, workforce availability, and patient demand influence wait times for medical procedures across the UK. Specifically, it examines the roles of hospital beds per capita, physicians per 1,000 inhabitants, and Type 1 attendance rates in shaping delays for procedures like knee replacements, coronary artery bypass grafts, and cataract surgeries. While previous studies have explored individual contributors to wait times, this analysis integrates multiple predictors to provide a holistic view of systemic pressures.

The findings reveal that declining hospital bed availability has a pronounced negative impact on wait times, while higher physician density and attendance rates are associated with additional delays, potentially due to demand outstripping capacity. These results emphasize the need for a balanced approach to resource allocation, combining infrastructure investment with demand management strategies.

Understanding the drivers of wait times is vital for improving healthcare access and efficiency. By identifying the factors contributing to delays, this paper offers actionable insights for policymakers and healthcare administrators seeking to enhance patient outcomes and system resilience. The remainder of this paper is structured as follows: Section 1 details the data and methodology, Section 2 presents key results, Section 3 discusses implications, and Section 4 concludes with recommendations for future research.

# Data {#sec-data}

## Overview

In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR], using the haven and `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al. 2021),. All figures in the report are generated using `ggplot2` [@citeGG]. We run the model in R [@citeR] using the `modelsummary` package of [@modelsummary].

This report integrates data from two key sources: the NHS and the OECD. These datasets provide critical information to analyze trends in healthcare demand and resources in the United Kingdom.

The NHS A&E Attendances and Emergency Admissions dataset captures the monthly demand for emergency services. It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission. These statistics, collected at the provider organization level, are drawn from NHS Trusts, Foundation Trusts, and independent sector organizations. The data was aggregated from monthly submissions, a change implemented after Sir Bruce Keogh’s review of waiting time standards. The dataset includes attendance and admission figures indexed to pre-pandemic and earlier baseline levels, enabling us to monitor trends and variations across years.

The remaining datasets are sourced from the OECD Health Statistics 2024 database, which provides standardized, internationally comparable data across healthcare systems. The OECD collaborates with organizations such as the United Nations and Eurostat to develop robust methodologies and benchmarks for health statistics. For the United Kingdom, data on hospital resources (e.g., hospital counts and bed availability), physician counts, and other health system metrics are provided by various entities like NHS Digital and Public Health Scotland. The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments.

## Measurement
**Physicians per 1,000** [@physicians]. Source: Physician counts are sourced from the OECD, relying on NHS Digital, Public Health Scotland, and the General Medical Council (GMC). The data reflect licensed physicians, encompassing both general practitioners and specialists. The metric is calculated as the total number of physicians per 1,000 population. The dataset includes both headcount and rolecount metrics, with post-2009 data transitioning to headcount for greater accuracy. Physician counts include GP retainers and full-time equivalents, providing an accurate picture of the available workforce. Historical data adjustments account for changes in collection methodologies. Similar datasets include WHO and Eurostat health statistics offer broader regional comparisons but lack the granularity needed for a UK-focused study.

**Beds per 1,000 Population** [@beds]: This metric, provided by the OECD, uses data from NHS Digital, Public Health Scotland, and national agencies across the UK. It tracks the availability of inpatient beds. Annual averages of beds available overnight in public hospitals. Includes acute care and psychiatric beds but excludes private sector facilities for consistency. Data are for financial years and represent publicly funded healthcare infrastructure. No significant methodological breaks are noted for this variable.

**Type 1 Major Care Attendances** [@attendance]: These figures represent percentages indexed to the baseline year 2011, providing a normalized measure of demand changes over time. The NHS uses administrative records to capture real-time data from provider organizations. The NHS employs administrative records to capture real-time data, validated through internal processes to mitigate potential inconsistencies or reporting delays. However, some unreported data at the trust level (e.g., support facilities and non-inpatient services) are excluded, which may lead to minor underestimations. Its about type 1 emergency admissions which are major

**Waiting Times for Key Medical Procedures** [@wait]: Waiting time data are sourced from the OECD, which uses NHS data for the United Kingdom. The dataset covers average waiting times for seven key procedures, including knee replacements and coronary artery bypass grafts. Procedure-Specific Wait Times: Measured in months, these reflect the average time in the NHS between referral and procedure. This dataset includes detailed procedure-specific insights, allowing for cross-procedure comparisons. Data validation is conducted at both the NHS and OECD levels to ensure reliability.

**Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors. While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not materially affect the study's conclusions. Similar datasets from organizations like WHO or Eurostat could offer supplementary insights but were not utilized due to a lack of UK-specific detail and consistency.

## Methodology #{sec-methodology}
EXPLAIN HOW YOU CLEANED DATA HERE

```{r}
#| echo: false
#| warning: false
#| error: false
#| label: tbl-data
#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
# Import the cleaned healthcare dataset
cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

# Split the table into smaller parts
part1 <- cleaned_data[, 1:6]  
part2 <- cleaned_data[, 7:ncol(cleaned_data)]

# Display the first part of the table
kable(head(part1), 
      caption = "Part 1: Hospital statistics")
```
```{r}
#| echo: false
#| warning: false
#| error: false
#| label: tbl-data-2
#| tbl-cap: "Wait times for medical procedures displayed separately"

# Update column names for the second part of the table
colnames(part2) <- gsub("^Wait_", "", colnames(part2))

# Display the second part of the table
kable(head(part2), 
      caption = "Part 2: Wait Times for Key Medical Procedures in Months")

```
@tbl-data presents the cleaned dataset presents hospital and healthcare statistics, consisting of 10 variables and 5 observations spanning the years 2015 to 2019. The variables include year, hospitals per 1,000,000 inhabitants, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, emergency admissions, and attendance rates. Additional variables capture wait times (in months) for key medical procedures such as coronary grafts, angioplasty, knee surgery, hip surgery, cataract surgery, hysterectomy, and prostatectomy. The average percentage increase in wait times was calculated relative to 2015. All metrics are standardized based on population or healthcare utilization data for each year.

## Outcome variables

```{r}
#| echo: false
#| warning: false
#| error: false
#| label: fig-data
#| tbl-cap: "Percentage Increase in Wait Times Compared to 2015 for Key Medical Procedures and Overall Average"
# Calculate percentage change for each medical procedure
cleaned_data <- cleaned_data |>
  mutate(across(
    starts_with("Wait_"), 
    ~ round((. / .[Year == 2015] - 1) * 100, 2),  # Calculate percentage change relative to 2015
    .names = "Pct_{.col}"  # Prefix the new columns with "Pct_"
  ))

# Reshape the percentage increase data to long format for plotting
long_wait_pct_data <- cleaned_data |>
  pivot_longer(
    cols = starts_with("Pct_Wait_"), 
    names_to = "Medical_Procedure",
    values_to = "Pct_Change"
  ) |>
  mutate(
    Medical_Procedure = ifelse(Medical_Procedure == "Pct_Wait_Avg", "Average Wait Time", 
                               gsub("Pct_Wait_", "", Medical_Procedure))  # Clean up procedure names
  )

# Plot all percentage increases on the same plot, including the average
ggplot(long_wait_pct_data, aes(x = Year, y = Pct_Change, color = Medical_Procedure, group = Medical_Procedure)) +
  geom_line(aes(linetype = Medical_Procedure), size = 1) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c(
      "Average Wait Time" = "black",
      setNames(rainbow(length(unique(long_wait_pct_data$Medical_Procedure)) - 1), 
               unique(long_wait_pct_data$Medical_Procedure)[unique(long_wait_pct_data$Medical_Procedure) != "Average Wait Time"])
    )
  ) +
  scale_linetype_manual(
    values = c(
      "Average Wait Time" = "dashed",
      setNames(rep("solid", length(unique(long_wait_pct_data$Medical_Procedure)) - 1), 
               unique(long_wait_pct_data$Medical_Procedure)[unique(long_wait_pct_data$Medical_Procedure) != "Average Wait Time"])
    )
  ) +
  labs(
    title = "Percentage Increase in Wait Times Compared to 2015",
    x = "Year",
    y = "Percentage Increase (%)",
    color = "Medical Procedure",
    linetype = "Medical Procedure"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```
@fig-data depicts the percentage increase in wait times for various medical procedures compared to 2015. By 2019, most procedures show a notable increase in wait times, with angioplasty and prostatectomy exhibiting the largest rises. While some procedures, such as knee and hip surgeries, show consistent increases, others, like prostatectomy, initially decline before rising sharply. The steady increase across most procedures highlights a systemic trend of growing delays in accessing medical care, potentially signaling resource limitations or increased demand in healthcare services. Overall, wait times have risen by over 20% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.
## Predictor variables

### Physicians per 1,000

```{r}
#| label: fig-physicians
#| fig-cap: United Kingdom's Physician rate per 1000 Inhabitants from 2015 to 2019
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = Year, y = physicians_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Physicians per 1 000 rates from 2015 to 2019", x = "Year", y = "Physicians per 1 000") +
  theme_minimal()
```
@fig-physicians shows a steady increase in the number of physicians per 1,000 inhabitants from 2015 to 2019, rising from approximately 2.0 to 2.15. This consistent growth indicates an improvement in physician availability relative to population size, potentially reflecting increased investment in healthcare workforce development or recruitment efforts. The trend suggests progress in addressing healthcare access, but further analysis is needed to determine whether this growth aligns with demand and regional healthcare needs.

### Hospital Beds per 1,000
```{r}
#| label: fig-beds
#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2019
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
  theme_minimal()
```
@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2019, decreasing from approximately 2.6 to 2.45 beds. This consistent downward trend suggests a reduction in bed capacity relative to population growth, potentially indicating healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure. The decline highlights the need to assess the impact on patient care and explore strategies to balance population needs with available resources.

### Demand for Services 
```{r}
#| label: fig-attendance-demand
#| fig-cap: "Type 1 Major Care Attendance in the United Kingdom from 2015 to 2019"
#| echo: false
#| warning: false

# Plotting Type 1 Major Care Attendance
ggplot(cleaned_data) +
  geom_line(aes(x = Year, y = attendance, color = "A&E Attendances"), size = 1) +
  scale_color_manual(values = c("A&E Attendances" = "blue")) +
  labs(
    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2019)",
    x = "Year",
    y = "Percentage Change from 2015 (%)",
    color = "Service Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```
@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2019. A&E attendances showed a modest, steady increase over the observed period, plateauing slightly by 2019. This trend highlights the growing demand for emergency services and underscores the need for further investigation into factors contributing to increased attendance, such as population growth, access to primary care, or changes in healthcare-seeking behavior. Addressing these factors through improved preventive and community-based care could help reduce the burden on emergency services.

# Model
After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., number of physicians, hospital beds per 1,000 people, and attendance rates) and wait times for medical procedures. These variables show potential predictive power, indicating a linear relationship. To further investigate and make predictions about how these factors influence wait times, we will implement a linear regression model.

The goal of this linear regression model is to estimate the coefficients $\beta_0, \beta_1, \beta_3$ such that the model fits the data well and provides insights into how each predictor contributes to wait times for medical procedures. Additionally, this model will enable predictions of wait times under different healthcare conditions.

The statistical significance of each $\beta_k$ coefficient will be assessed using a t-test, testing whether the coefficient is significantly different from zero. If the p-value for a coefficient is less than a chosen significance level (e.g., 0.05), we can conclude that the corresponding predictor has a significant effect on wait times.

## Model set-up
This model framework allows us to evaluate the influence of healthcare system factors (like hospital and physician availability) on wait times for specific procedures. Understanding these relationships is critical for designing policies aimed at reducing wait times and improving healthcare delivery.

We expect a positive relationship between the average percentage increase in wait times (Pct_Wait_Time) and healthcare system pressures, such as a decline in the number of hospital beds per 1,000 inhabitants or hospitals per million people. Conversely, we anticipate a negative relationship between Pct_Wait_Time and the availability of physicians per 1,000 inhabitants, as greater physician availability could lead to faster patient care and reduced wait times.

The final model is displayed here:

$$
Y_{i} = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \beta_3X_{3i} + \epsilon_{i}
$$

\begin{align} 
$Y_{ij}$ is the average percenta change in wait time for the 7 specific medical procedures in year $i$.\\
$X_{1i}, X_{2i}, X_{3i}$ represent the independent variables of physicians per 1000 people, beds per 1000 people, Type 1 attendance for procedure $j$ in year $i$.\\
$\beta_0$ is the intercept or constant term, representing the expected wait time when all predictors are zero.\\
$\beta_1, \beta_2, \beta_3$ are the slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.\\
$\epsilon_{i}$ is the error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.
\end{align}

Each predictor variable is expected to contribute uniquely:

- Beds per 1,000 inhabitants (X_1): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
- Physicians per 1,000 inhabitants (X_2): Greater physician availability is hypothesized to decrease wait times.
- Attendance Rates (X_3): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.

We run the model in R [@citeR] using the `modelsummary` package of...

### Model justification

This model framework allows us to evaluate the impact of key healthcare system factors on wait times. The selected predictors—beds_per_1000, physicians_per_1000, and attendance—are directly linked to resource availability and demand, making them relevant for understanding delays in medical care. The results from this model can provide actionable insights to inform policies aimed at optimizing resource allocation, improving healthcare infrastructure, and managing patient flow to reduce wait times.

We expect to confirm that reduced healthcare resources (fewer hospital beds, lower physician availability), and increased demand (more tpe 1 attendances) are associated with longer wait times, while better resource availability can mitigate delays. This analysis will help identify key leverage points for optimizing resource allocation and reducing wait times across various medical procedures.

# Results

## Model Validation

```{r}
#| echo: false
#| output: asis
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors"

# Fit the linear regression model
model <- lm(
   Pct_Wait_Avg ~ beds_per_1000 + physicians_per_1000 + attendance,
  data = cleaned_data
)

# Display the model summary in a table
modelsummary::modelsummary(
  list("Linear Model" = model),
  statistic = c("p.value")
)

```

@tbl-modelresults presents the linear regression model which reveals significant relationships between healthcare system factors and the percentage increase in wait times as all the variables included in the model have p-values less than 0.05. The number of hospital beds per 1,000 inhabitants has a strong negative effect (\Beta=−75.36), indicating that as bed availability decreases, wait times increase. Physician availability per 1,000 shows a positive but smaller effect (\Beta=29.99), suggesting that higher physician counts may correlate with increased demand or delays. Attendance rates also have a significant positive effect (\Beta=1.45), highlighting the strain of growing patient loads on healthcare systems. The model's $R^2$ of 1 suggests it explains nearly all variability in the data, though this may reflect overfitting given the small sample size. Overall, the results emphasize the critical role of resource availability in managing wait times.

# Discussion

## Overview
This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for key medical procedures in the United Kingdom from 2015 to 2019. Our analysis employs linear modeling to find that the availability of hospital beds per 1,000 inhabitants plays a crucial role in mitigating delays, with reduced bed capacity significantly increasing wait times. While higher physician density was hypothesized to reduce delays, our findings suggest a complex relationship where increased physician numbers can exacerbate delays due to system inefficiencies or heightened demand. Additionally, rising attendance rates place further strain on healthcare resources, leading to longer wait times. These findings emphasize the multifaceted nature of healthcare delays and underscore the need for targeted policy interventions to address both resource allocation and demand management.

### Relationship between Wait Times and Beds per 1,000

The results demonstrate a significant negative relationship between the number of hospital beds per 1,000 inhabitants and wait times, emphasizing the critical role of inpatient capacity in managing healthcare demands. As bed availability decreases, hospitals face greater difficulty accommodating patient volumes, leading to longer delays in elective and urgent procedures. This finding aligns with existing healthcare literature, which consistently highlights the importance of adequate infrastructure to ensure timely care delivery. The steep coefficient suggests that even small reductions in bed availability can have outsized effects, creating ripple effects throughout the healthcare system. For example, limited bed availability can delay the discharge of patients in acute care settings, reducing the ability to admit new patients. This underscores the need for careful resource management and forward-looking infrastructure planning, especially in the context of aging populations and rising healthcare demands. Long-term solutions could include increasing investment in hospital infrastructure, optimizing the use of existing beds, and integrating innovative models of care, such as home-based hospitalization, to reduce the strain on inpatient services.

### Relationship between Wait Times and Physicians per 1,000

The observed positive relationship between physician availability and wait times appears counterintuitive but is explainable within the broader healthcare context. Regions with higher physician densities may experience increased diagnostic and referral rates, as greater access to physicians can result in more patients entering the healthcare system. This could lead to heightened demand for specialized procedures, thereby increasing average wait times. Additionally, inefficiencies in resource allocation, such as misaligned staffing or insufficient operating room availability, may prevent the full utilization of increased physician capacity to reduce delays. The results suggest that increasing physician numbers alone is insufficient to address wait times without concurrent improvements in systemic efficiency. Policymakers should focus on streamlining workflows, enhancing interdisciplinary collaboration, and ensuring that physician availability is complemented by adequate infrastructure and support staff. These measures can help prevent bottlenecks and maximize the impact of physician density on reducing delays.

### Relationship between Wait Times and Type 1 Major Attendance

The positive relationship between Type 1 major attendance rates and wait times underscores the systemic strain caused by rising patient loads. As attendance rates increase, hospitals face escalating pressure to prioritize emergency cases, which can divert resources from elective and non-urgent procedures. This phenomenon is particularly concerning in systems like the NHS, where universal healthcare ensures that emergency care is always prioritized. Increased attendance may be driven by a range of factors, including population growth, limited access to primary care, and changing healthcare-seeking behaviors. Without interventions to address these root causes, hospitals are likely to continue struggling with resource allocation, leading to prolonged wait times across the board. Potential strategies to mitigate this strain include expanding primary care access to reduce unnecessary emergency department visits, implementing community-based triaging systems, and improving public health education to guide appropriate healthcare utilization. Addressing attendance pressures holistically could help alleviate the burden on emergency departments and reduce delays for elective procedures.

## Limitations 
One major limitation of this analysis is the small sample size, which includes only five observations spanning 2015 to 2019. This limits the generalizability of the findings and raises concerns about potential overfitting in the regression model. The limited time frame also fails to capture longer-term trends or cyclical variations, such as seasonal fluctuations in demand for specific procedures. Furthermore, the dataset lacks information on critical confounding variables, such as regional healthcare policies, socioeconomic disparities, or detailed patient demographics, which could significantly influence wait times. For instance, regions with different funding allocations or staffing strategies may exhibit divergent wait time trends for which this study cannot account.

## Future Work
To address the limitations of this study, future research should prioritize expanding the dataset to include more years and a broader range of healthcare metrics. Incorporating variables such as regional funding levels and technological investments could provide a more comprehensive understanding of wait time determinants. Additionally, future studies could explore non-linear relationships or interaction effects, such as how the interplay between physician density and hospital bed availability influences delays. Applying machine learning approaches to identify latent patterns or leveraging simulation models to predict the impact of policy interventions could further enhance the analytical depth of this research. By addressing these avenues, future work can better inform targeted policies to optimize resource allocation, reduce wait times, and improve healthcare delivery.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

\newpage

# References

### 2024-11-24 scripts/03-clean_data.R

@@ -1,111 +1,147 @@
 #### Preamble ####
-# Purpose: Cleans the raw healthcare data from OECD.
+# Purpose: Cleans and merges raw healthcare data from OECD into one dataset with physicians normalized per 1000 population.
 # Author: Chenika Bukes
 # Date: 20 November 2024
 # Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
-# Any other information needed? None
-
-
-# Load the library
-# library(arrow)
-
-# Read the CSV file
-df <- read.csv("./data/01-raw_data/hospital_beds.csv")
-
-# Save the data frame as a Parquet file
-parquet_file_path <- "./data/01-raw_data/hospital_beds.parquet"
-write_parquet(df, parquet_file_path)
-
 
 #### Workspace setup ####
 library(arrow)
 library(tidyverse)
 
-#### Clean data ####
-# Load the raw Parquet file
-raw_data <- read_parquet("./data/01-raw_data/waiting_times.parquet")
 
-# Filter for required conditions and select relevant columns
-cleaned_wait_times <- raw_data |>
+#### Process and merge datasets ####
+
+# 1. Load and process the waiting times data
+waiting_times_data <- read_parquet("./data/01-raw_data/waiting_times.parquet") |>
   filter(Reference.area == "United Kingdom" & 
            TIME_PERIOD >= 2015 & 
            TIME_PERIOD <= 2019) |>
   select(TIME_PERIOD, OBS_VALUE, Medical.procedure) |>
   rename(
+    Year = TIME_PERIOD,
     WaitTime = OBS_VALUE
   ) |>
-  arrange(Year)
-
-# Load the raw Parquet file
-raw_data <- read_parquet("./data/01-raw_data/number_hospitals.parquet")
-
-# Filter for required conditions and select relevant columns
-cleaned_hospital_data <- raw_data |>
-  filter(Reference.area == "United Kingdom" & 
-           TIME_PERIOD >= 2015 & 
-           TIME_PERIOD <= 2019 &
-           Hospital.type == "Total") |>
-  select(TIME_PERIOD, OBS_VALUE) |>
-  rename(
-    Year = TIME_PERIOD,
-    HospitalsPerMillionInhabitants = OBS_VALUE
+  pivot_wider(
+    names_from = Medical.procedure,
+    values_from = WaitTime,
+    names_prefix = "WaitTime_"
   ) |>
   arrange(Year)
 
[…]

### 2024-11-24 paper/paper.qmd
Added data citation and package citations

@@
 format: pdf
+toc: true
 number-sections: true
@@
 By identifying the factors contributing to delays, this paper offers actionable insights for policymakers and healthcare administrators seeking to enhance patient outcomes and system resilience.
-The remainder of this paper is structured as follows: Section 1 details the data and methodology, Section 2 presents key results, Section 3 discusses implications, and Section 4 concludes with recommendations for future research.
+The remainder of this paper is structured as follows: @sec-data details the data and methodology, @sec-model presents the model used for analysis of the data, @sec-results presents the key results from the model, and @sec-discussion concludes with final analysis of the results and recommendations for future research.
 
@@
 In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR], using the haven and `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al.
-2021),.
+2021).
 All figures in the report are generated using `ggplot2` [@citeGG].
@@
 
-The NHS A&E Attendances and Emergency Admissions dataset captures the monthly demand for emergency services.
+The NHS A&E Attendances and Emergency Admissions dataset [@attendance] captures the monthly demand for emergency services.
 It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission.
@@
 
-The remaining datasets are sourced from the OECD Health Statistics 2024 database, which provides standardized, internationally comparable data across healthcare systems.
+The remaining datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data across healthcare systems.
 The OECD collaborates with organizations such as the United Nations and Eurostat to develop robust methodologies and benchmarks for health statistics.
@@
 **Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors.
-While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not materially affect the study's conclusions.
+While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2019.
 Similar datasets from organizations like WHO or Eurostat could offer supplementary insights but were not utilized due to a lack of UK-specific detail and consistency.
 
-## Methodology #{sec-methodology}
+## Methodology {#sec-methodology}
 EXPLAIN HOW YOU CLEANED DATA HERE
@@
 
+### Hospital Beds per 1,000
+```{r}
+#| label: fig-beds
+#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2019
+#| echo: false
+#| warning: false
+
+ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
+  geom_line(color = "blue") +
+  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
+  theme_minimal()
+```
+
+@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2019, decreasing from approximately 2.6 to 2.45 beds.
+This consistent downward trend suggests a reduction in bed capacity relative to population growth, potentially indicating healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure.
+The decline highlights the need to assess the impact on patient care and explore strategies to balance population needs with available resources.
+
 ### Physicians per 1,000
@@
 ```
+
 @fig-physicians shows a steady increase in the number of physicians per 1,000 inhabitants from 2015 to 2019, rising from approximately 2.0 to 2.15.
@@
 The trend suggests progress in addressing healthcare access, but further analysis is needed to determine whether this growth aligns with demand and regional healthcare needs.
-
-### Hospital Beds per 1,000
-```{r}
-#| label: fig-beds
-#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2019
-#| echo: false
-#| warning: false
-
-ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
-  geom_line(color = "blue") +
-  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
-  theme_minimal()
-```
-@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2019, decreasing from approximately 2.6 to 2.45 beds.
-This consistent downward trend suggests a reduction in bed capacity relative to population growth, potentially indicating healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure.
-The decline highlights the need to assess the impact on patient care and explore strategies to balance population needs with available resources.
 
@@
   theme(legend.position = "bottom")
-
 ```
@@
 
-# Model
+# Model {#sec-model}
 After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., number of physicians, hospital beds per 1,000 people, and attendance rates) and wait times for medical procedures.
@@
 \begin{align} 
-$Y_{ij}$ is the average percenta change in wait time for the 7 specific medical procedures in year $i$.\\
-$X_{1i}, X_{2i}, X_{3i}$ represent the independent variables of physicians per 1000 people, beds per 1000 people, Type 1 attendance for procedure $j$ in year $i$.\\
-$\beta_0$ is the intercept or constant term, representing the expected wait time when all predictors are zero.\\
-$\beta_1, \beta_2, \beta_3$ are the slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.\\
-$\epsilon_{i}$ is the error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.
+Y_{ij} & = \text{The average percentage change in wait time for the 7 specific medical procedures in year } i. \\
+X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
+\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \\
+\beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
+\epsilon_{i} & = \text{The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.}
 \end{align}
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-11-24 paper/paper.qmd
Completed model file and imported model to main paper.

@@
 
-Understanding the drivers of wait times is vital for improving healthcare access and efficiency.
+Understanding the drivers of wait times is vital for improving healthcare access and efficiency [@king].
 By identifying the factors contributing to delays, this paper offers actionable insights for policymakers and healthcare administrators seeking to enhance patient outcomes and system resilience.
@@
 **Physicians per 1,000** [@physicians].
-Source: Physician counts are sourced from the OECD, relying on NHS Digital, Public Health Scotland, and the General Medical Council (GMC).
+Physician counts are sourced from the OECD, relying on NHS Digital, Public Health Scotland, and the General Medical Council (GMC).
 The data reflect licensed physicians, encompassing both general practitioners and specialists.
@@
 # Split the table into smaller parts
-part1 <- cleaned_data[, 1:6]  
-part2 <- cleaned_data[, 7:ncol(cleaned_data)]
+part1 <- cleaned_data[, 1:4]  
+part2 <- cleaned_data[, 4:ncol(cleaned_data)]
 
@@
 
-# Fit the linear regression model
-model <- lm(
-   Pct_Wait_Avg ~ beds_per_1000 + physicians_per_1000 + attendance,
-  data = cleaned_data
-)
+# Load the saved model
+model <- readRDS("../models/wait_times_model.rds")
 
@@
 modelsummary::modelsummary(
-  list("Linear Model" = model),
+  list("Frequentist Linear Model" = model),
   statistic = c("p.value")
 )
-
 ```

### 2024-11-24 scripts/03-clean_data.R

@@ -10,7 +10,6 @@
 library(arrow)
 library(tidyverse)
 
-
 #### Process and merge datasets ####
 
 # 1. Load and process the waiting times data

### 2024-11-24 scripts/06-model_data.R

@@ -1,37 +1,25 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Fits and saves a frequentist linear model for hospital wait times.
+# Author: Chenika Bukes
+# Date: 24 November 2024
+# Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: Cleaned dataset is available at "./data/02-analysis_data/final_healthcare_data.csv".
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+# Load the cleaned data
+analysis_data <- read_csv("./data/02-analysis_data/final_healthcare_data.csv")
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
-
-
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+#### Fit the linear model ####
+# Model: Percentage change in wait times as a function of healthcare predictors
+wait_times_model <- lm(
+  Pct_Wait_Avg ~ beds_per_1000 + physicians_per_1000 + attendance,
+  data = analysis_data
 )
 
-
+#### Save the model ####
+# Save the fitted model for reuse
+saveRDS(wait_times_model, file = "./models/wait_times_model.rds")

### 2024-11-24 scripts/00-simulate_data.R
Completed simulations and tests of simulated and analysis data.

@@ -1,52 +1,58 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulates a dataset of healthcare statistics similar to the described dataset.
+# Author: Chenika Bukes
+# Date: 24 November 2024
+# Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Pre-requisites: The `tidyverse` package must be installed.
 
 #### Workspace setup ####
 library(tidyverse)
-set.seed(853)
-
+set.seed(66)
 
 #### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
+# Years for the data
+years <- 2015:2019
+
+# Generate data for each year
+simulated_data <- tibble(
+  Year = rep(years, each = 5),  # 5 observations per year
+  beds_per_1000 = rnorm(n = 25, mean = 2.5, sd = 0.2),  
+  physicians_per_1000 = rnorm(n = 25, mean = 2.07, sd = 0.1),  
+  attendance = rnorm(n = 25, mean = 100, sd = 15),  
+  
+  # Simulated wait times for procedures (in months)
+  Wait_CorGraft = rnorm(n = 25, mean = 6, sd = 1.5),
+  Wait_Angioplasty = rnorm(n = 25, mean = 5, sd = 1.2),
+  Wait_Knee = rnorm(n = 25, mean = 8, sd = 2),
+  Wait_Hip = rnorm(n = 25, mean = 7, sd = 1.8),
+  Wait_Cataract = rnorm(n = 25, mean = 3, sd = 0.8),
+  Wait_Prostatectomy = rnorm(n = 25, mean = 4, sd = 1.1),
+  Wait_Hysterectomy = rnorm(n = 25, mean = 5.5, sd = 1.3)
 )
 
-# Political parties
-parties <- c("Labor", "Liberal", "Greens", "National", "Other")
+# Calculate percentage increases relative to 2015 for each procedure
+base_year <- simulated_data %>%
[…]

### 2024-11-24 scripts/01-test_simulated_data.R

@@ -1,23 +1,22 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the simulated healthcare dataset.
+# Author: Chenika Bukes
+# Date: 24 November 2024
+# Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# - The `tidyverse` package must be installed and loaded
+# - The simulated healthcare dataset must exist
+# Any other information needed? Ensure that the dataset is generated and saved in the appropriate directory.
 
 
 #### Workspace setup ####
 library(tidyverse)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+simulated_data <- read_csv("data/00-simulated_data/simulated_healthcare_data.csv")
 
 # Test if the data was successfully loaded
-if (exists("analysis_data")) {
+if (exists("simulated_data")) {
   message("Test Passed: The dataset was successfully loaded.")
 } else {
   stop("Test Failed: The dataset could not be loaded.")
@@ -26,64 +25,84 @@ if (exists("analysis_data")) {
 
 #### Test data ####
 
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
+# Check if the dataset has 25 rows (5 years, 5 observations per year)
+if (nrow(simulated_data) == 25) {
+  message("Test Passed: The dataset has 25 rows.")
 } else {
-  stop("Test Failed: The dataset does not have 151 rows.")
+  stop("Test Failed: The dataset does not have 25 rows.")
 }
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
+# Check if the dataset has the correct number of columns
+expected_columns <- c(
+  "Year", "beds_per_1000", "physicians_per_1000", "attendance",
+  "Wait_CorGraft", "Wait_Angioplasty", "Wait_Knee", "Wait_Hip",
+  "Wait_Cataract", "Wait_Prostatectomy", "Wait_Hysterectomy",
[…]

### 2024-11-24 scripts/03-test_analysis_data.R

@@ -1,0 +1,66 @@
+#### Preamble ####
+# Purpose: Tests the structure and validity of the cleaned healthcare dataset.
+# Author: Chenika Bukes
+# Date: 24 November 2024
+# Contact: chenika.bukes@mail.utoronto.ca
+# License: MIT
+# Pre-requisites:
+#   - The `tidyverse` and `testthat` packages must be installed and loaded
+#   - The cleaned dataset must exist at "data/02-analysis_data/final_healthcare_data.csv"
+
+#### Workspace setup ####
+library(tidyverse)
+library(testthat)
+
+cleaned_data <- read_csv("../data/02-analysis_data/final_healthcare_data.csv")
+
+#### Test data ####
+
+# Test that the dataset has 5 rows (one per year, 2015–2019)
+test_that("dataset has 5 rows", {
+  expect_equal(nrow(cleaned_data), 5)
+})
+
+# Test that all values in 'Year' are integers between 2015 and 2019
+test_that("'Year' column contains valid years", {
+  expect_true(all(cleaned_data$Year %in% 2015:2019))
+})
+
+# Test that 'beds_per_1000' and 'physicians_per_1000' are within realistic ranges
+test_that("'beds_per_1000' and 'physicians_per_1000' are within realistic ranges", {
+  expect_true(all(cleaned_data$beds_per_1000 >= 1 & cleaned_data$beds_per_1000 <= 3))
+  expect_true(all(cleaned_data$physicians_per_1000 >= 1.5 & cleaned_data$physicians_per_1000 <= 2.5))
+})
+
+# Test that 'attendance' is non-negative
+test_that("'attendance' column is non-negative", {
+  expect_true(all(cleaned_data$attendance >= 0))
+})
+
+# Test that all waiting times are non-negative
+test_that("waiting time columns are non-negative", {
+  waiting_time_columns <- grep("^Wait_", names(cleaned_data), value = TRUE)
+  expect_true(all(sapply(cleaned_data[waiting_time_columns], function(col) all(col >= 0, na.rm = TRUE))))
+})
+
+# Test that 'Pct_Wait_Avg' is non-negative
+test_that("'Pct_Wait_Avg' is non-negative", {
+  expect_true(all(cleaned_data$Pct_Wait_Avg >= 0))
+})
+
+# Test that there are no missing values in the dataset
+test_that("no missing values in dataset", {
[…]

### 2024-11-24 scripts/00-simulate_data.R
Using testthat in the simulated data tests.

@@ -14,27 +14,36 @@ set.seed(66)
 # Years for the data
 years <- 2015:2019
 
+# Function to ensure non-negative random values
+generate_non_negative <- function(n, mean, sd, min_value = 0) {
+  abs(rnorm(n, mean, sd))  # Ensure all values are non-negative
+}
+
 # Generate data for each year
 simulated_data <- tibble(
   Year = rep(years, each = 5),  # 5 observations per year
-  beds_per_1000 = rnorm(n = 25, mean = 2.5, sd = 0.2),  
-  physicians_per_1000 = rnorm(n = 25, mean = 2.07, sd = 0.1),  
-  attendance = rnorm(n = 25, mean = 100, sd = 15),  
+  beds_per_1000 = generate_non_negative(n = 25, mean = 2.5, sd = 0.2),  
+  physicians_per_1000 = generate_non_negative(n = 25, mean = 2.07, sd = 0.1),  
+  attendance = generate_non_negative(n = 25, mean = 100, sd = 15),  
   
   # Simulated wait times for procedures (in months)
-  Wait_CorGraft = rnorm(n = 25, mean = 6, sd = 1.5),
-  Wait_Angioplasty = rnorm(n = 25, mean = 5, sd = 1.2),
-  Wait_Knee = rnorm(n = 25, mean = 8, sd = 2),
-  Wait_Hip = rnorm(n = 25, mean = 7, sd = 1.8),
-  Wait_Cataract = rnorm(n = 25, mean = 3, sd = 0.8),
-  Wait_Prostatectomy = rnorm(n = 25, mean = 4, sd = 1.1),
-  Wait_Hysterectomy = rnorm(n = 25, mean = 5.5, sd = 1.3)
+  Wait_CorGraft = generate_non_negative(n = 25, mean = 6, sd = 1.5),
+  Wait_Angioplasty = generate_non_negative(n = 25, mean = 5, sd = 1.2),
+  Wait_Knee = generate_non_negative(n = 25, mean = 8, sd = 2),
+  Wait_Hip = generate_non_negative(n = 25, mean = 7, sd = 1.8),
+  Wait_Cataract = generate_non_negative(n = 25, mean = 3, sd = 0.8),
+  Wait_Prostatectomy = generate_non_negative(n = 25, mean = 4, sd = 1.1),
+  Wait_Hysterectomy = generate_non_negative(n = 25, mean = 5.5, sd = 1.3)
 )
 
[…]

### 2024-11-24 scripts/01-test_simulated_data.R

@@ -5,34 +5,30 @@
 # Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-# - The `tidyverse` package must be installed and loaded
-# - The simulated healthcare dataset must exist
+#   - The `tidyverse` and `testthat` packages must be installed and loaded
+#   - The simulated healthcare dataset must exist
 # Any other information needed? Ensure that the dataset is generated and saved in the appropriate directory.
 
 
 #### Workspace setup ####
 library(tidyverse)
+library(testthat)
 
-simulated_data <- read_csv("data/00-simulated_data/simulated_healthcare_data.csv")
-
-# Test if the data was successfully loaded
-if (exists("simulated_data")) {
-  message("Test Passed: The dataset was successfully loaded.")
-} else {
-  stop("Test Failed: The dataset could not be loaded.")
-}
-
+simulated_data <- read_csv("../data/00-simulated_data/simulated_healthcare_data.csv")
 
 #### Test data ####
 
-# Check if the dataset has 25 rows (5 years, 5 observations per year)
-if (nrow(simulated_data) == 25) {
-  message("Test Passed: The dataset has 25 rows.")
-} else {
-  stop("Test Failed: The dataset does not have 25 rows.")
-}
+# Test that the dataset was successfully loaded
+test_that("Dataset is loaded", {
+  expect_true(exists("simulated_data"))
+})
 
-# Check if the dataset has the correct number of columns
+# Test that the dataset has 25 rows (5 years, 5 observations per year)
+test_that("Dataset has 25 rows", {
+  expect_equal(nrow(simulated_data), 25)
+})
+
+# Test that the dataset has the expected columns
 expected_columns <- c(
   "Year", "beds_per_1000", "physicians_per_1000", "attendance",
   "Wait_CorGraft", "Wait_Angioplasty", "Wait_Knee", "Wait_Hip",
@@ -41,68 +37,51 @@ expected_columns <- c(
   "Pct_Wait_Hip", "Pct_Wait_Cataract", "Pct_Wait_Prostatectomy", 
   "Pct_Wait_Hysterectomy", "Pct_Wait_Avg"
 )
+test_that("Dataset has correct columns", {
+  expect_equal(colnames(simulated_data), expected_columns)
+})
 
-if (all(colnames(simulated_data) == expected_columns)) {
-  message("Test Passed: The dataset contains the correct columns.")
-} else {
[…]

### 2024-11-26 paper/paper.qmd
Updated data to include years 2019-2022.

@@
 library(modelsummary)
+library(patchwork)
 ```
@@
 While previous studies have explored individual contributors to wait times, this analysis integrates multiple predictors to provide a holistic view of systemic pressures.
+
+The estimand in this study is the average percentage change in wait times for seven key medical procedures: knee replacements, hip replacements, prostatectomy, hysterectomy, coronary artery bypass grafts, transluminal coronary angioplasty, and cataract surgeries in the UK.
+The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
+This measure quantifies how changes in healthcare infrastructure and demand impact system-wide delays, offering insights into the relative contribution of each predictor to observed wait times.
 
@@
 2021).
-All figures in the report are generated using `ggplot2` [@citeGG].
+All figures in the report are generated using `ggplot2` [@citeGG] and `patchwork` [@patchwork].
 We run the model in R [@citeR] using the `modelsummary` package of [@modelsummary].
@@
 
+The NHS Referral to Treatment (RTT) waiting times dataset [@wait] is the primary source for waiting time statistics.
+These data measure the median time (in weeks) between referral by a general practitioner (GP) and the start of treatment for elective care, capturing delays across 15 key medical procedures such as general surgery, urology, and trauma & orthopedics.
+This dataset offers procedure-specific insights that allow for cross-procedure and temporal comparisons.
+
 The NHS A&E Attendances and Emergency Admissions dataset [@attendance] captures the monthly demand for emergency services.
 It includes the total number of attendances at Accident & Emergency (A&E) departments and emergency admissions, alongside measures of wait times for admission.
-These statistics, collected at the provider organization level, are drawn from NHS Trusts, Foundation Trusts, and independent sector organizations.
-The data was aggregated from monthly submissions, a change implemented after Sir Bruce Keogh’s review of waiting time standards.
-The dataset includes attendance and admission figures indexed to pre-pandemic and earlier baseline levels, enabling us to monitor trends and variations across years.
+These statistics are collected at the provider organization level from NHS Trusts, Foundation Trusts, and independent sector organizations.
+The data are aggregated from monthly submissions, a system adopted following Sir Bruce Keogh’s review of waiting time standards, and include attendance and admission figures.
 
-The remaining datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data across healthcare systems.
-The OECD collaborates with organizations such as the United Nations and Eurostat to develop robust methodologies and benchmarks for health statistics.
-For the United Kingdom, data on hospital resources (e.g., hospital counts and bed availability), physician counts, and other health system metrics are provided by various entities like NHS Digital and Public Health Scotland.
-The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments.
+Additional datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data on healthcare systems.
+For the United Kingdom, the OECD compiles data on hospital resources hospital beds and physician counts from NHS Digital and Public Health Scotland.
+The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments, making it a reliable complement to NHS data for the analysis.
 
@@
 Historical data adjustments account for changes in collection methodologies.
+The data are collected and reported monthly through NHS administrative systems, with automated validation checks ensuring accuracy and consistency.
 Similar datasets include WHO and Eurostat health statistics offer broader regional comparisons but lack the granularity needed for a UK-focused study.
+[@https://turasdata.nes.nhs.scot/about-our-data-and-reports/methods/nhsscotland-workforce-data-warehouse/]
 
@@
 **Type 1 Major Care Attendances** [@attendance]: These figures represent percentages indexed to the baseline year 2011, providing a normalized measure of demand changes over time.
-The NHS uses administrative records to capture real-time data from provider organizations.
 The NHS employs administrative records to capture real-time data, validated through internal processes to mitigate potential inconsistencies or reporting delays.
 However, some unreported data at the trust level (e.g., support facilities and non-inpatient services) are excluded, which may lead to minor underestimations.
-Its about type 1 emergency admissions which are major
+Type 1 Major Care is defined as major emergency departments that provide a consultant-led 24-hour service with full facilities for resuscitating patients [@care].
 
-**Waiting Times for Key Medical Procedures** [@wait]: Waiting time data are sourced from the OECD, which uses NHS data for the United Kingdom.
-The dataset covers average waiting times for seven key procedures, including knee replacements and coronary artery bypass grafts.
-Procedure-Specific Wait Times: Measured in months, these reflect the average time in the NHS between referral and procedure.
-This dataset includes detailed procedure-specific insights, allowing for cross-procedure comparisons.
-Data validation is conducted at both the NHS and OECD levels to ensure reliability.
+**Waiting Times for Key Medical Procedures** [@wait]: Waiting time data are sourced directly from NHS England’s Referral to Treatment (RTT) waiting times statistics, as published on NHS England’s statistics portal.
+These statistics measure the time elapsed between a patient’s referral by a general practitioner (GP) and the initiation of treatment, providing insights into NHS performance.
+Waiting times are calculated as the median duration (in weeks) for patients to begin their procedures, ensuring consistency across treatment categories.
+The dataset includes procedure-specific insights for key medical interventions such as general surgery, urology, and trauma & orthopedics, among others, enabling cross-procedure comparisons.
+Data are collected monthly through NHS administrative systems and validated using automated checks and standardized reporting protocols to ensure accuracy and consistency.
 
@@
 While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2019.
-Similar datasets from organizations like WHO or Eurostat could offer supplementary insights but were not utilized due to a lack of UK-specific detail and consistency.
+Similar datasets from the World Health Organization (WHO) could offer supplementary insights but were not utilized due to a lack of UK-specific detail.
 
 ## Methodology {#sec-methodology}
-EXPLAIN HOW YOU CLEANED DATA HERE
+The data cleaning method is described in @sec-cleaning.
+All data cleaning was done using `arrow` [@arrow], `tidyverse` [@tidyverse], `readxl` [@readxl], and `stringr` [@stringr].
+Below is a snapshot of the first five rows of the key data columns which will be used for this paper.
+@tbl-data-appendix provides the entire cleaned dataset in table format.
 
@@
 # Split the table into smaller parts
-part1 <- cleaned_data[, 1:4]  
-part2 <- cleaned_data[, 4:ncol(cleaned_data)]
+part1 <- cleaned_data[, c(1:4, which(colnames(cleaned_data) == "Total"))]
 
@@
 ```
-```{r}
-#| echo: false
-#| warning: false
-#| error: false
-#| label: tbl-data-2
-#| tbl-cap: "Wait times for medical procedures displayed separately"
 
-# Update column names for the second part of the table
-colnames(part2) <- gsub("^Wait_", "", colnames(part2))
-
-# Display the second part of the table
-kable(head(part2), 
-      caption = "Part 2: Wait Times for Key Medical Procedures in Months")
-
-```
-@tbl-data presents the cleaned dataset presents hospital and healthcare statistics, consisting of 10 variables and 5 observations spanning the years 2015 to 2019.
-The variables include year, hospitals per 1,000,000 inhabitants, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, emergency admissions, and attendance rates.
-Additional variables capture wait times (in months) for key medical procedures such as coronary grafts, angioplasty, knee surgery, hip surgery, cataract surgery, hysterectomy, and prostatectomy.
-The average percentage increase in wait times was calculated relative to 2015.
+@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of 10 variables and 8 observations spanning the years 2015 to 2022.
+The variables include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
+Additional variables capture wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery.
+In @tbl-data we see the average percentage change in wait times from 2015 levels in the `Total` column.
 All metrics are standardized based on population or healthcare utilization data for each year.
@@
 ## Outcome variables
+
+@fig-data depicts the percentage increase in wait times for various medical procedures compared to 2015.
+By 2019, most procedures show a notable increase in wait times, with angioplasty and prostatectomy exhibiting the largest rises.
+While some procedures, such as knee and hip surgeries, show consistent increases, others, like prostatectomy, initially decline before rising sharply.
+The steady increase across most procedures highlights a systemic trend of growing delays in accessing medical care, potentially signaling resource limitations or increased demand in healthcare services.
+Overall, wait times have risen by over 20% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.
 
@@
 #| error: false
-#| label: fig-data
-#| tbl-cap: "Percentage Increase in Wait Times Compared to 2015 for Key Medical Procedures and Overall Average"
-# Calculate percentage change for each medical procedure
-cleaned_data <- cleaned_data |>
-  mutate(across(
-    starts_with("Wait_"), 
-    ~ round((. / .[Year == 2015] - 1) * 100, 2),  # Calculate percentage change relative to 2015
-    .names = "Pct_{.col}"  # Prefix the new columns with "Pct_"
-  ))
+#| label: fig-total
+#| tbl-cap: "Percentage Increase in Total Wait Times Compared to 2015"
 
-# Reshape the percentage increase data to long format for plotting
-long_wait_pct_data <- cleaned_data |>
-  pivot_longer(
-    cols = starts_with("Pct_Wait_"), 
-    names_to = "Medical_Procedure",
-    values_to = "Pct_Change"
-  ) |>
-  mutate(
-    Medical_Procedure = ifelse(Medical_Procedure == "Pct_Wait_Avg", "Average Wait Time", 
-                               gsub("Pct_Wait_", "", Medical_Procedure))  # Clean up procedure names
+# Plot for the "Total" column
+ggplot(long_wait_pct_data %>% filter(Treatment == "Total"), 
+       aes(x = Year, y = Pct_Change, group = Treatment)) +
+  geom_line(size = 1, color = "black") +
+  geom_point(size = 2, color = "black") +
+  labs(
+    title = "Percentage Change in Total Wait Times Compared to 2015",
+    x = "Year",
+    y = "Percentage Change (%)",
+    color = NULL
+  ) +
+  theme_minimal() +
+  theme(
+    legend.position = "none",
+    text = element_text(size = 14)
   )
 
-# Plot all percentage increases on the same plot, including the average
-ggplot(long_wait_pct_data, aes(x = Year, y = Pct_Change, color = Medical_Procedure, group = Medical_Procedure)) +
-  geom_line(aes(linetype = Medical_Procedure), size = 1) +
-  geom_point(size = 2) +
-  scale_color_manual(
-    values = c(
-      "Average Wait Time" = "black",
-      setNames(rainbow(length(unique(long_wait_pct_data$Medical_Procedure)) - 1), 
-               unique(long_wait_pct_data$Medical_Procedure)[unique(long_wait_pct_data$Medical_Procedure) != "Average Wait Time"])
-    )
-  ) +
-  scale_linetype_manual(
-    values = c(
-      "Average Wait Time" = "dashed",
-      setNames(rep("solid", length(unique(long_wait_pct_data$Medical_Procedure)) - 1), 
-               unique(long_wait_pct_data$Medical_Procedure)[unique(long_wait_pct_data$Medical_Procedure) != "Average Wait Time"])
-    )
-  ) +
-  labs(
-    title = "Percentage Increase in Wait Times Compared to 2015",
-    x = "Year",
-    y = "Percentage Increase (%)",
-    color = "Medical Procedure",
-    linetype = "Medical Procedure"
-  ) +
-  theme_minimal() +
-  theme(legend.position = "bottom")
+```
 
-```
-@fig-data depicts the percentage increase in wait times for various medical procedures compared to 2015.
-By 2019, most procedures show a notable increase in wait times, with angioplasty and prostatectomy exhibiting the largest rises.
-While some procedures, such as knee and hip surgeries, show consistent increases, others, like prostatectomy, initially decline before rising sharply.
-The steady increase across most procedures highlights a systemic trend of growing delays in accessing medical care, potentially signaling resource limitations or increased demand in healthcare services.
-Overall, wait times have risen by over 20% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.
+@fig-total shows the percentage increase in total wait times for key medical procedures in the UK relative to 2015, revealing a steady rise through 2019, followed by fluctuations during the COVID-19 pandemic.
+The moderate growth between 2015 and 2019 suggests manageable strain on healthcare resources, but the dip in 2020 reflects reduced elective procedures as hospitals prioritized emergency care.
+This was followed by a sharp increase in 2021 and a steep surge in 2022, likely due to the backlog created during the pandemic, ongoing demand pressures, and resource constraints.
+These trends underscore the urgent need for targeted healthcare interventions, including expanded capacity, streamlined workflows, and policies to address the post-pandemic backlog and increasing demand.
 ## Predictor variables
@@
 
-@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2019, decreasing from approximately 2.6 to 2.45 beds.
-This consistent downward trend suggests a reduction in bed capacity relative to population growth, potentially indicating healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure.
-The decline highlights the need to assess the impact on patient care and explore strategies to balance population needs with available resources.
+@fig-beds shows a steady decline in hospital beds per 1,000 inhabitants from 2015 to 2020, decreasing from approximately 2.6 to 2.45 beds.
+This consistent downward trend indicates a reduction in bed capacity relative to population growth, reflecting potential healthcare system pressures such as resource reallocation, efficiency measures, or underinvestment in infrastructure.
+However, a slight rebound is observed in 2021 and 2022, with the number of beds per 1,000 inhabitants stabilizing at just above 2.45.
+This recent stabilization could suggest efforts to address capacity constraints or temporary adjustments in response to the increased demand for hospital resources during the COVID-19 pandemic.
+The trend underscores the importance of evaluating the impact of long-term capacity changes on patient care and exploring strategies to ensure sufficient resources to meet population needs.
 
@@
 
-@fig-physicians shows a steady increase in the number of physicians per 1,000 inhabitants from 2015 to 2019, rising from approximately 2.0 to 2.15.
-This consistent growth indicates an improvement in physician availability relative to population size, potentially reflecting increased investment in healthcare workforce development or recruitment efforts.
-The trend suggests progress in addressing healthcare access, but further analysis is needed to determine whether this growth aligns with demand and regional healthcare needs.
+@fig-physicians shows a steady increase in the number of physicians per 1,000 inhabitants from 2015 to 2022, rising from approximately 2.8 to 3.2.
+This consistent growth reflects an improvement in physician availability relative to population size, likely driven by increased investment in healthcare workforce development, recruitment efforts, and possibly adjustments in healthcare policies.
+The marked acceleration between 2019 and 2021 may reflect efforts to bolster healthcare capacity during the COVID-19 pandemic.
+While this trend highlights progress in enhancing workforce availability, further analysis is needed to assess whether these increases adequately meet rising healthcare demands and address regional disparities in physician distribution.
 
@@
 ```
-@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2019.
-A&E attendances showed a modest, steady increase over the observed period, plateauing slightly by 2019.
-This trend highlights the growing demand for emergency services and underscores the need for further investigation into factors contributing to increased attendance, such as population growth, access to primary care, or changes in healthcare-seeking behavior.
-Addressing these factors through improved preventive and community-based care could help reduce the burden on emergency services.
+@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2022.
+Between 2015 and 2019, A&E attendances showed a modest, steady increase, reflecting a growing demand for emergency services.
+However, there was a sharp decline in 2020, likely attributable to the COVID-19 pandemic and related restrictions, which disrupted normal healthcare-seeking behavior.
+Attendance levels rebounded significantly in 2021, surpassing pre-pandemic levels, and remained stable through 2022.
+This trend highlights both the resilience of demand for emergency services and the need for strategies to manage surges in attendance effectively, such as enhancing access to primary care and preventive healthcare services.
 
 # Model {#sec-model}
-After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., number of physicians, hospital beds per 1,000 people, and attendance rates) and wait times for medical procedures.
-These variables show potential predictive power, indicating a linear relationship.
-To further investigate and make predictions about how these factors influence wait times, we will implement a linear regression model.
+After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., physicians per 1,000 people, hospital beds per 1,000 people, and Type 1 attendance rates) and wait times for medical procedures.
+These variables indicate potential predictive power, suggesting a linear relationship.
+To further investigate and quantify how these factors influence wait times, we implemented a linear regression model.
 
@@
 The statistical significance of each $\beta_k$ coefficient will be assessed using a t-test, testing whether the coefficient is significantly different from zero.
-If the p-value for a coefficient is less than a chosen significance level (e.g., 0.05), we can conclude that the corresponding predictor has a significant effect on wait times.
+If the p-value for a coefficient is less than a chosen significance level, we can conclude that the corresponding predictor has a significant effect on wait times.
 
 ## Model set-up
-This model framework allows us to evaluate the influence of healthcare system factors (like hospital and physician availability) on wait times for specific procedures.
-Understanding these relationships is critical for designing policies aimed at reducing wait times and improving healthcare delivery.
-
-We expect a positive relationship between the average percentage increase in wait times (Pct_Wait_Time) and healthcare system pressures, such as a decline in the number of hospital beds per 1,000 inhabitants or hospitals per million people.
-Conversely, we anticipate a negative relationship between Pct_Wait_Time and the availability of physicians per 1,000 inhabitants, as greater physician availability could lead to faster patient care and reduced wait times.
-
 The final model is displayed here:
@@
 \begin{align} 
-Y_{ij} & = \text{The average percentage change in wait time for the 7 specific medical procedures in year } i. \\
+Y_{ij} & = \text{The average percentage change in wait time for the 15 specific medical procedures in year i}. \\
 X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
-\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \\
+\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \
 \beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
@@
 \end{align}
+
+The goal of the linear regression model is to estimate the values of \beta_1, \beta_2, and \beta_3 such that the model fits the data well and to predict the expected value of the average total waiting time for procedures for different values of physician rates, hospital bed rates, and type 1 major care attendance.
+The statistical significance of 1 can be assessed using a t-test, which tests whether the estimated coeﬀicient is significantly different from zero.
+If the p-value of the t-test for each coefficient is less than a chosen significance level, we can conclude that there is a significant relationship between the average percentage change in wait time and the healthcare predictors.
 
@@
 
-We run the model in R [@citeR] using the `modelsummary` package of...
+We run the model in R [@citeR] using the `modelsummary` package of `rsaratnam` [@rsaratnam].
 
@@
 # Results {#sec-results}
-
-## Model Validation
 
@@
 
-@tbl-modelresults presents the linear regression model which reveals significant relationships between healthcare system factors and the percentage increase in wait times as all the variables included in the model have p-values less than 0.05.
-The number of hospital beds per 1,000 inhabitants has a strong negative effect (\Beta=−75.36), indicating that as bed availability decreases, wait times increase.
-Physician availability per 1,000 shows a positive but smaller effect (\Beta=29.99), suggesting that higher physician counts may correlate with increased demand or delays.
-Attendance rates also have a significant positive effect (\Beta=1.45), highlighting the strain of growing patient loads on healthcare systems.
-The model's $R^2$ of 1 suggests it explains nearly all variability in the data, though this may reflect overfitting given the small sample size.
-Overall, the results emphasize the critical role of resource availability in managing wait times.
+The linear regression model, presented in @tbl-modelresults, explores the relationships between healthcare system factors and the percentage change in wait times, though none of the variables demonstrate statistically significant effects, as indicated by p-values greater than 0.05.
+The number of hospital beds per 1,000 inhabitants shows a small positive effect (\beta=4.71), suggesting a minimal increase in wait times with higher bed availability, though this effect lacks statistical significance.
+Physician availability per 1,000 (\beta=31.83) indicates a larger positive association, yet it is also not statistically significant.
+Attendance rates (\beta =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times.
+The model achieves an $R^2$ value of 0.599, indicating that approximately 60% of the variability in wait times is explained by the model, though the adjusted $R^2$ value of 0.299 reflects limited predictive power.
+With a sample size of n=8, the findings should be interpreted cautiously, as the lack of significance and low adjusted $R^2$ highlight potential limitations in the model's robustness and generalizability.
+
+```{r}
+#| echo: false
+#| output: asis
+#| eval: true
+#| label: fig-modelresults1
+#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining percentage increase in wait times based on healthcare system predictors."
+
+# Load the saved model
+model <- readRDS("../models/wait_times_model.rds")
+
+# Load the dataset used for the model
+data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+
+# Function to create a projection plot for a specific predictor
+create_projection_plot <- function(predictor_name, data, model) {
+  # Create a grid of values for the predictor of interest
+  grid <- seq(min(data[[predictor_name]], na.rm = TRUE), 
+              max(data[[predictor_name]], na.rm = TRUE), length.out = 100)
+
+  # Create a new data frame for prediction
+  pred_data <- data.frame(grid)
+  colnames(pred_data) <- predictor_name
+
+  # Fix all other predictors at their mean values
+  other_predictors <- setdiff(names(model$coefficients)[-1], predictor_name)
+  for (var in other_predictors) {
+    pred_data[[var]] <- mean(data[[var]], na.rm = TRUE)
+  }
+
+  # Predict using the model
+  pred_data$Predicted <- predict(model, newdata = pred_data)
+
+  # Create the plot
+  ggplot(data, aes_string(x = predictor_name, y = "Total")) +
+    geom_point(color = "blue", size = 2, alpha = 0.7) +
+    geom_line(data = pred_data, aes_string(x = predictor_name, y = "Predicted"), 
+              color = "red", size = 1) +
+    labs(
+      x = predictor_name,
+      y = "Wait Times (%)"
+    ) +
+    theme_minimal() +
+    theme(aspect.ratio = 0.8)
+}
+
+# Generate individual plots
+p1 <- create_projection_plot("attendance", data, model)
+p2 <- create_projection_plot("physicians_per_1000", data, model)
+p3 <- create_projection_plot("beds_per_1000", data, model)
+
+# Combine the plots into one figure
+combined_plot <- (p1 | p2 | p3) +
+  plot_annotation(
+    title = "Projections of Linear Model onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
+    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
+  )
+
+# Display the combined plot
+print(combined_plot)
+
+```
+The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (Pct_Wait_Avg) compared to three key predictors: attendance, physicians_per_1000, and beds_per_1000.
+The red lines depict the predicted relationships, while the blue points represent the observed data values.
+The attendance predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend.
+Physicians_per_1000 also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance.
+In contrast, beds_per_1000 shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend.
+These plots suggest that attendance is the most influential predictor of wait times among the three variables, while beds_per_1000 has the least impact.
+
+@fig-modeldetails presents the residual plots.
+The residual plots indicate that the linear model generally fits the data well, with residuals scattered randomly around zero for all predictors, suggesting no major violations of linearity or independence assumptions.
+However, in each residual plot there is a violation of constant variance where there are noticeable outliers and potential influential points, particularly in the plots for attendance and physicians per 1,000, which may affect the model's robustness.
+The fitted vs.
+actual values plot shows a good overall alignment with most points near the predicted line, but two significant deviations highlight discrepancies that could stem from these outliers.
+Given the small sample size, these findings should be interpreted with caution.
 
@@
 ## Overview
-This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for key medical procedures in the United Kingdom from 2015 to 2019.
-Our analysis employs linear modeling to find that the availability of hospital beds per 1,000 inhabitants plays a crucial role in mitigating delays, with reduced bed capacity significantly increasing wait times.
-While higher physician density was hypothesized to reduce delays, our findings suggest a complex relationship where increased physician numbers can exacerbate delays due to system inefficiencies or heightened demand.
-Additionally, rising attendance rates place further strain on healthcare resources, leading to longer wait times.
-These findings emphasize the multifaceted nature of healthcare delays and underscore the need for targeted policy interventions to address both resource allocation and demand management.
+This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for key medical procedures in the United Kingdom from 2015 to 2022.
+Our analysis employs a linear regression model but finds no statistically significant relationships between the predictors and wait times.
+While hospital beds per 1,000 inhabitants showed a weak positive association with wait times, the relationship was not significant, suggesting that bed capacity alone may not directly mitigate delays.
+Similarly, physician density and attendance rates demonstrated positive but insignificant associations with wait times, reflecting the complex interplay of factors influencing healthcare delays.
+These findings indicate that systemic inefficiencies, resource allocation challenges, and unmeasured variables may play a more significant role than initially hypothesized.
+The limited explanatory power of the model and the presence of anomalies in the data post-2019 highlight the need for more detailed data and refined analyses to uncover the true drivers of healthcare delays.
 
@@
 
-The results demonstrate a significant negative relationship between the number of hospital beds per 1,000 inhabitants and wait times, emphasizing the critical role of inpatient capacity in managing healthcare demands.
-As bed availability decreases, hospitals face greater difficulty accommodating patient volumes, leading to longer delays in elective and urgent procedures.
-This finding aligns with existing healthcare literature, which consistently highlights the importance of adequate infrastructure to ensure timely care delivery.
-The steep coefficient suggests that even small reductions in bed availability can have outsized effects, creating ripple effects throughout the healthcare system.
-For example, limited bed availability can delay the discharge of patients in acute care settings, reducing the ability to admit new patients.
-This underscores the need for careful resource management and forward-looking infrastructure planning, especially in the context of aging populations and rising healthcare demands.
-Long-term solutions could include increasing investment in hospital infrastructure, optimizing the use of existing beds, and integrating innovative models of care, such as home-based hospitalization, to reduce the strain on inpatient services.
+The weak positive association between hospital beds per 1,000 inhabitants and wait times was unexpected and not statistically significant.
+While prior research emphasizes the importance of bed availability in managing healthcare demands, this study suggests that increasing bed numbers alone may not guarantee reduced delays.
+The minimal effect size indicates that factors such as bed utilization efficiency, patient flow management, and the quality of care delivery could overshadow the raw number of beds available.
+For instance, hospitals with higher bed capacity but poor discharge planning or limited staffing may still experience significant delays.
+These findings underscore the importance of complementing infrastructure investments with systemic reforms, such as optimizing bed turnover rates, enhancing care coordination, and adopting alternative care models like outpatient procedures or home-based care.
 
@@
 
-The observed positive relationship between physician availability and wait times appears counterintuitive but is explainable within the broader healthcare context.
-Regions with higher physician densities may experience increased diagnostic and referral rates, as greater access to physicians can result in more patients entering the healthcare system.
-This could lead to heightened demand for specialized procedures, thereby increasing average wait times.
-Additionally, inefficiencies in resource allocation, such as misaligned staffing or insufficient operating room availability, may prevent the full utilization of increased physician capacity to reduce delays.
-The results suggest that increasing physician numbers alone is insufficient to address wait times without concurrent improvements in systemic efficiency.
-Policymakers should focus on streamlining workflows, enhancing interdisciplinary collaboration, and ensuring that physician availability is complemented by adequate infrastructure and support staff.
-These measures can help prevent bottlenecks and maximize the impact of physician density on reducing delays.
+The positive but statistically insignificant relationship between physician density and wait times challenges the expectation that greater workforce availability reduces delays.
+This result may reflect systemic inefficiencies or increased demand driven by greater access to healthcare services.
+For example, higher physician availability can lead to more diagnostic and treatment referrals, overwhelming other parts of the healthcare system, such as specialized care or operating rooms.
+Additionally, unmeasured factors like disparities in physician specialization, workflow inefficiencies, or misalignment between physician density and patient needs could further obscure the relationship.
+These findings highlight that increasing physician numbers alone is insufficient to reduce wait times without addressing systemic bottlenecks.
+Interventions such as improving care pathways, enhancing interdisciplinary collaboration, and aligning physician availability with infrastructure and support staff could better leverage workforce capacity to reduce delays.
 
@@
 
-The positive relationship between Type 1 major attendance rates and wait times underscores the systemic strain caused by rising patient loads.
-As attendance rates increase, hospitals face escalating pressure to prioritize emergency cases, which can divert resources from elective and non-urgent procedures.
-This phenomenon is particularly concerning in systems like the NHS, where universal healthcare ensures that emergency care is always prioritized.
-Increased attendance may be driven by a range of factors, including population growth, limited access to primary care, and changing healthcare-seeking behaviors.
-Without interventions to address these root causes, hospitals are likely to continue struggling with resource allocation, leading to prolonged wait times across the board.
-Potential strategies to mitigate this strain include expanding primary care access to reduce unnecessary emergency department visits, implementing community-based triaging systems, and improving public health education to guide appropriate healthcare utilization.
-Addressing attendance pressures holistically could help alleviate the burden on emergency departments and reduce delays for elective procedures.
+The weak positive association between attendance rates and wait times, though not significant, aligns with broader concerns about rising patient loads straining healthcare systems.
+As attendance increases, hospitals may prioritize emergency cases, diverting resources from elective and non-urgent procedures, potentially exacerbating delays.
+Factors such as limited access to primary care, changes in population demographics, and healthcare-seeking behaviors likely drive the observed trends.
+Although the relationship was not statistically significant, addressing the root causes of rising attendance remains critical to reducing systemic strain.
+Potential strategies include expanding primary care access, implementing community triaging systems, and improving public health education to promote appropriate healthcare utilization.
+These measures could help balance patient loads and reduce delays for non-emergency procedures.
 
 ## Limitations 
-One major limitation of this analysis is the small sample size, which includes only five observations spanning 2015 to 2019.
-This limits the generalizability of the findings and raises concerns about potential overfitting in the regression model.
-The limited time frame also fails to capture longer-term trends or cyclical variations, such as seasonal fluctuations in demand for specific procedures.
-Furthermore, the dataset lacks information on critical confounding variables, such as regional healthcare policies, socioeconomic disparities, or detailed patient demographics, which could significantly influence wait times.
-For instance, regions with different funding allocations or staffing strategies may exhibit divergent wait time trends for which this study cannot account.
+The primary limitation of this study is the small sample size, which includes only eight observations spanning 2015 to 2022.
+This limits the generalizability of findings and weakens the statistical power needed to detect significant relationships.
+Additionally, the dataset is influenced by the effects of COVID-19, with wait times deviating significantly from the pre-pandemic trend.
+As illustrated in @fig-wait, wait times followed a consistent linear increase from 2015 to 2019, but 2020 deviated sharply below trend, while 2021 exhibited a sharp increase above trend.
+These anomalies suggest inconsistent variance in the data and indicate that a linear model may not be appropriate for post-2019 data.
+Moreover, the dataset lacks information on key confounding variables such as regional healthcare policies, socioeconomic disparities, or detailed patient demographics, which could significantly influence wait times.
+Unmeasured systemic inefficiencies, such as staffing misalignment, operating room availability, or referral practices, may also account for the insignificant results.
+Furthermore, the linear model does not account for potential interaction effects or non-linear relationships between predictors, which could offer deeper insights into the dynamics of healthcare delays.
 
 ## Future Work
-To address the limitations of this study, future research should prioritize expanding the dataset to include more years and a broader range of healthcare metrics.
-Incorporating variables such as regional funding levels and technological investments could provide a more comprehensive understanding of wait time determinants.
-Additionally, future studies could explore non-linear relationships or interaction effects, such as how the interplay between physician density and hospital bed availability influences delays.
-Applying machine learning approaches to identify latent patterns or leveraging simulation models to predict the impact of policy interventions could further enhance the analytical depth of this research.
-By addressing these avenues, future work can better inform targeted policies to optimize resource allocation, reduce wait times, and improve healthcare delivery.
+To address the limitations of this study, future research should prioritize collecting monthly data for the years 2015 to 2019 to explore more granular trends before the anomalies introduced by COVID-19.
+Analyzing these pre-pandemic patterns could provide a clearer understanding of the role that metrics such as beds per 1,000, physician density, and attendance rates played during a relatively stable period.
+Additionally, expanding the dataset to include broader healthcare metrics, such as regional funding levels, technological investments, and patient demographics, could yield a more comprehensive analysis.
+Future studies should also explore non-linear models or interaction effects to capture the intricate relationships between predictors.
+Simulation models and scenario analyses could be used to test policy interventions, such as optimizing resource allocation or improving workflow efficiency, to identify actionable strategies for reducing wait times and improving healthcare delivery.
 
@@
 
-# Additional data details
+## Full Analysis Dataset
 
-# Model details {#sec-model-details}
+```{r}
+#| echo: false
+#| warning: false
+#| error: false
+#| label: tbl-data-appendix
+#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
+# Import the cleaned healthcare dataset
+cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+
+# Split the table into smaller parts
+part1 <- cleaned_data[, c(1:6)]
+part2 <- cleaned_data[, c(6:12)]
+part3 <- cleaned_data[, c(12:18)]
+part4 <- cleaned_data[, c(18:ncol(cleaned_data))]
+
+# Display the first part of the table
+kable(head(part1), 
+      caption = "Part 1: Hospital statistics")
+kable(head(part2), 
+      caption = "Part 2: Hospital statistics")
+kable(head(part3), 
+      caption = "Part 3: Hospital statistics")
+kable(head(part4), 
+      caption = "Part 4: Hospital statistics")
+```
+
+## Data Cleaning {#sec-cleaning}
+The data cleaning process undertaken in this project involved several key steps to ensure the dataset was accurate, standardized, and ready for analysis.
+First, raw data on healthcare waiting times, physicians, hospital beds, and emergency admissions from various sources, such as the OECD and NHS, were loaded.
+For the waiting times data, all Excel files were processed by standardizing column names, extracting relevant fields (treatment function and average waiting time), and appending metadata such as the year and month derived from file names.
+Treatment functions were harmonized across years, with those from April 2021 onward mapped to their pre-2021 equivalents using a predefined renaming scheme.
+Data with irrelevant or missing treatment functions (e.g., "Other") were excluded.
+
+Yearly averages for each treatment function were calculated by grouping the data by year and function and computing the mean waiting time.
+The data was then pivoted to make years the rows and treatment functions the columns, providing a wide-format structure suitable for further analysis.
+Special corrections were applied, such as manually imputing specific values for ENT in 2019 and 2020.
+
+All datasets were merged on the year variable to produce a consolidated dataset.
+Percent changes in waiting times from 2015 levels were computed for each treatment function to highlight trends over time.
+Lastly, columns with multiple-word treatment names were shortened to their first word, with duplicates (e.g., "General Surgery" and "General Medicine") resolved by appending unique suffixes.
+
+### Full Outcome Variable Plots
+```{r}
+#| echo: false
+#| warning: false
+#| error: false
+#| label: fig-data-group1
+#| tbl-cap: "Percentage Increase in Wait Times (Group 1)"
+#| fig-height: 8
+#| fig-width: 12
+
+# Reshape the data to long format for plotting
+long_wait_pct_data <- cleaned_data |>
+  pivot_longer(
+    cols = -Year, # Include all columns except 'Year'
+    names_to = "Treatment",
+    values_to = "Pct_Change"
+  ) |>
+  filter(!Treatment %in% c("attendance", "physicians_per_1000", "beds_per_1000", "General_S"))
+
+# Split treatments into three groups and ensure "Total" is included in all
+treatments <- unique(long_wait_pct_data$Treatment)
+group_size <- ceiling((length(treatments) - 1) / 3)  # Exclude "Total" when dividing groups
+group1 <- c("Total", treatments[1:group_size])
+group2 <- c("Total", treatments[(group_size + 1):(2 * group_size)])
+group3 <- c("Total", treatments[(2 * group_size + 1):length(treatments)])
+
+# Define colors for treatments, ensuring "Total" is black
+generate_colors <- function(group) {
+  setNames(
+    c("black", rainbow(length(group) - 1)),
+    group
+  )
+}
+
+# Plot for the first group
+colors_group1 <- generate_colors(group1)
+
+plot1 <- long_wait_pct_data %>%
+  filter(Treatment %in% group1) %>%
+  ggplot(aes(x = Year, y = Pct_Change, color = Treatment, group = Treatment)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  scale_color_manual(
+    values = colors_group1,
+    guide = guide_legend(title = "Treatment")
+  ) +
+  labs(
+    title = "Percentage Increase in Wait Times (Group 1)",
+    x = "Year",
+    y = "Percentage Increase (%)"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "bottom", text = element_text(size = 14))
+
+print(plot1)
+
+```
+```{r}
+#| echo: false
+#| warning: false
+#| error: false
+#| label: fig-data-group2
+#| tbl-cap: "Percentage Increase in Wait Times (Group 2)"
+#| fig-height: 8
+#| fig-width: 12
+
+# Plot for the first group
+colors_group2 <- generate_colors(group1)
+
+plot1 <- long_wait_pct_data %>%
+  filter(Treatment %in% group2) %>%
+  ggplot(aes(x = Year, y = Pct_Change, color = Treatment, group = Treatment)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  scale_color_manual(
+    values = colors_group1,
+    guide = guide_legend(title = "Treatment")
+  ) +
+  labs(
+    title = "Percentage Increase in Wait Times (Group 2)",
+    x = "Year",
+    y = "Percentage Increase (%)"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "bottom", text = element_text(size = 14))
+
+print(plot2)
+
+```
+```{r}
+#| echo: false
+#| warning: false
+#| error: false
+#| label: fig-data-group2
+#| tbl-cap: "Percentage Increase in Wait Times (Group 3)"
+#| fig-height: 8
+#| fig-width: 12
+
+# Plot for the second group
+plot3 <- long_wait_pct_data %>%
+  filter(Treatment %in% group3) %>%
+  ggplot(aes(x = Year, y = Pct_Change, color = Treatment, group = Treatment)) +
+  geom_line(size = 1) +
+  geom_point(size = 2) +
+  labs(
+    title = "Percentage Increase in Wait Times (Group 3)",
+    x = "Year",
+    y = "Percentage Increase (%)",
+    color = "Treatment"
+  ) +
+  theme_minimal() +
+  theme(legend.position = "bottom", text = element_text(size = 14))
+
+print(plot3)
+```
+## Model Details {#sec-model-details}
+### Original Model Residuals
+```{r}
+#| echo: false
+#| output: asis
+#| eval: true
+#| label: fig-modeldetails
+#| tbl-cap: "Residual plots of each predictor."
+#| # Load necessary libraries
+library(ggplot2)
+
+# Extract residuals and fitted values from the model
+data$residuals <- residuals(model)
+data$fitted <- fitted.values(model)
+
+# Plot b: Residuals vs Predictor 1 (attendance)
+plot_b <- ggplot(data, aes(x = attendance, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Attendance",
+    y = "Residuals",
+    x = "Attendance"
+  )
+
+# Plot c: Residuals vs Predictor 2 (physicians_per_1000)
+plot_c <- ggplot(data, aes(x = physicians_per_1000, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Physicians per 1,000",
+    y = "Residuals",
+    x = "Physicians per 1,000"
+  )
+
+# Plot d: Fitted values vs Actual values (Pct_Wait_Avg)
+plot_e <- ggplot(data, aes(x = Total, y = fitted)) +
+  geom_point(color = "blue") +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
+  theme_classic() +
+  labs(
+    title = "Fitted vs Actual Values",
+    y = "Fitted Values",
+    x = "Actual Values"
+  )
+
+# Plot e: Residuals vs Predictor 3 (beds_per_1000)
+plot_d <- ggplot(data, aes(x = beds_per_1000, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Beds per 1,000",
+    y = "Residuals",
+    x = "Beds per 1,000"
+  )
+
+# Combine all plots into a grid
+combined_plot <- (plot_b | plot_c) / (plot_d | plot_e ) +
+  plot_annotation(title = "Diagnostic Plots for Linear Model")
+
+# Print the combined plot
+print(combined_plot)
+```
+### Adapted Model 
+```{r}
+#| echo: false
+#| output: asis
+#| eval: true
+#| label: tbl-modeladj
+#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors"
+
+#### Workspace setup ####
+library(tidyverse)
+
+#### Read data ####
+# Load the cleaned data
+analysis_data <- read_csv("../data/02-analysis_data/final_healthcare_data.csv")
+
+#### Filter data ####
+# Exclude data from 2020 to 2022
+filtered_data <- analysis_data %>% 
+  filter(Year < 2020)
+
+#### Fit the linear model ####
+# Model: Percentage change in wait times as a function of healthcare predictors
+wait_times_model_adj <- lm(
+  Total ~ beds_per_1000 + physicians_per_1000 + attendance,
+  data = filtered_data
+)
+
+#### Save the model ####
+# Save the fitted model for reuse
+saveRDS(wait_times_model_adj, file = "../models/wait_times_model_pre_2020.rds")
+
+# Load the saved model
+modela <- readRDS("../models/wait_times_model_pre_2020.rds")
+
+# Display the model summary in a table
+modelsummary::modelsummary(
+  list("Frequentist Linear Model" = modela),
+  statistic = c("p.value")
+)
+
+```
 

### 2024-11-26 scripts/02-clean_data.R

@@ -8,60 +8,162 @@
 
 #### Workspace setup ####
 library(arrow)
+library(readxl)
 library(tidyverse)
+library(stringr)
+library(dplyr)
 
 #### Process and merge datasets ####
 
 # 1. Load and process the waiting times data
-waiting_times_data <- read_parquet("./data/01-raw_data/waiting_times.parquet") |>
-  filter(Reference.area == "United Kingdom" & 
-           TIME_PERIOD >= 2015 & 
-           TIME_PERIOD <= 2019) |>
-  select(TIME_PERIOD, OBS_VALUE, Medical.procedure) |>
-  rename(
-    Year = TIME_PERIOD,
-    WaitTime = OBS_VALUE
-  ) |>
-  pivot_wider(
-    names_from = Medical.procedure,
-    values_from = WaitTime,
-    names_prefix = "WaitTime_"
-  ) |>
-  arrange(Year)
+# Define the path to the folder containing the Excel files for waiting times
+file_path <- "./data/01-raw_data"  
+file_list <- list.files(file_path, pattern = "*.xls", full.names = TRUE)
 
-# Calculate percentage increase for each operation relative to 2015
-wait_times_percentage <- waiting_times_data |>
-  mutate(across(
-    starts_with("WaitTime_"),
-    ~ round((. / .[Year == 2015] - 1) * 100, 2),  # Calculate percentage increase and round to 2 decimals
-    .names = "Pct_{.col}"               
-  ))
+# Define treatment function names (before and after April 2021)
+treatment_names_before_2021 <- c(
+  "General Surgery", "Urology", "Trauma & Orthopaedics", "ENT", "Ophthalmology", 
+  "Oral Surgery", "Neurosurgery", "Plastic Surgery", "Cardiothoracic Surgery", 
+  "General Medicine", "Gastroenterology", "Cardiology", "Dermatology", 
+  "Thoracic Medicine", "Neurology", "Rheumatology", "Geriatric Medicine", 
+  "Gynaecology", "Other", "Total"
+)
 
-# Add percentage increase columns for each operation relative to 2015
-waiting_times_percentage <- waiting_times_data |>
-  mutate(across(
-    starts_with("WaitTime_"),
-    ~ round((. / .[Year == 2015] - 1) * 100, 2),  # Calculate percentage increase and round to 2 decimals
-    .names = "Pct_{.col}"  # Add prefix "Pct_" to new columns
-  ))
+treatment_names_after_2021 <- c(
[…]

### 2024-11-26 scripts/05-model_data.R

@@ -16,10 +16,11 @@ analysis_data <- read_csv("./data/02-analysis_data/final_healthcare_data.csv")
 #### Fit the linear model ####
 # Model: Percentage change in wait times as a function of healthcare predictors
 wait_times_model <- lm(
-  Pct_Wait_Avg ~ beds_per_1000 + physicians_per_1000 + attendance,
+  Total ~ beds_per_1000 + physicians_per_1000 + attendance,
   data = analysis_data
 )
 
 #### Save the model ####
 # Save the fitted model for reuse
 saveRDS(wait_times_model, file = "./models/wait_times_model.rds")
+

### 2024-11-26 other/datasheet/datasheet_RTT.qmd
Wrote datasheet for RTT data.

---
title: "Datasheet for NHS RTT Waiting Times Dataset"
subtitle: "Understanding Patient Access to Treatment in England"
author: 
  - Chenika Bukes
date: today
date-format: long
format: pdf
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to monitor patient access to treatment within the NHS in England and support compliance with the 18-week waiting time target. It provides insights for healthcare planning, resource allocation, and policy development.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The dataset was created and maintained by NHS England's statistical team under the Statistical Work Areas Division.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - The dataset is funded and published by NHS England, as part of its public healthcare management responsibilities.
4. *Any other comments?*
    - This dataset is crucial for evaluating the efficiency of NHS services and identifying areas requiring intervention.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- The instances represent monthly statistics on NHS patient waiting times for treatment in England. They include data on the number of patients waiting for treatment and the number who received treatment within the reporting period.
2. *How many instances are there in total (of each type, if appropriate)?*
  - There are monthly instances of data on RTT waiting times from August 2007 until October 2024.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- The dataset includes all NHS Trusts and Clinical Commissioning Groups (CGG) in England, making it comprehensive.
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Each instance consists of structured data fields for 15 treatment types (e.g. cardiology, neurology) which include the number of patients waiting, the number treated within specific timeframes (e.g., within 1 week, 2 weeks, ..., up to 52 weeks), and the median waiting times in weeks for admission. These fields provide a detailed breakdown of patient wait times and treatment durations.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- No.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- Almost all months had reporting from all NHS locations. In October, November, and December 2022, data was missing from Frimley Health NHS Foundation Trust (RDU) and Manchester University NHS Foundation Trust (R0A). The reason for the missing data was not provided, but a supplemental time series instance was provided to account for the expected numbers from these NHS locations.  
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- Not applicable.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- No.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- No.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
  - Instances for years 2007 to 2010 are located on the national archive site which is publicly available. 
	- Instances for years 2010 to 2024 the dataset is self contained on the NHS website.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- No, all data is anonymized and aggregated to prevent identification of individuals.
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- No, the dataset contains operational statistics.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- No.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No, all data is aggregated and anonymized.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- No.
16. *Any other comments?*
	- Not Applicable.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- The data was collected via NHS Trusts and Clinical Commissioning Groups (CCGs), submitted monthly to NHS England.
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- NHS Trusts used internal systems to track referral-to-treatment pathways and submitted the data through standard reporting templates.
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- Not Applicable.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- NHS staff at Trusts and CCGs submitted the data. RTT data collection was conducted by Unify2 until March 2018 and then by the NHS Digital’s Strategic Data Collection Service. 
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- Data collection began in August 2007 and continues monthly.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Ethical review was not required as the dataset contains aggregated, anonymized data.
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- The data was collected directly by NHS Trusts and CCGs.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- No.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- No.
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- No.
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Yes, NHS England adheres to GDPR and data protection standards.
12. *Any other comments?*
	- The dataset is a vital tool for improving healthcare delivery and policy.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- Yes, the data was validated and cleaned by NHS England prior to publication. This process involved filtering treatment categories to ensure consistency and accuracy. The median waiting time was calculated, and the data was aggregated by waiting times in weekly intervals. Each column represents the number of patients who were treated within a specific number of weeks (e.g., 1 week, 2 weeks, ..., 52 weeks) during the reporting month, providing a structured and comprehensive view of patient wait times.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- The raw data is not publicly available.
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- No.
4. *Any other comments?*
	- Not applicable.

**Uses**
1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- Yes, it is widely used for monitoring NHS performance and research.
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- No
3. *What (other) tasks could the dataset be used for?*
	- Resource allocation, patient flow modeling, and healthcare policy analysis.
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- No.
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- It should not be used to infer individual patient outcomes.
6. *Any other comments?*
	- Not applicable.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- Yes, it is publicly available.
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- Via the NHS England website for public use [@wait].
	- Via secure email to ministers and those cleared to attain pre-releases of the data [@prerelease].
3. *When will the dataset be distributed?*
	- It is updated on the website monthly.
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- Yes, distributed under Open Government License.
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- No
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- No
7. *Any other comments?*
	- Ministers and selected officials receive pre-release access to official statistics once they are in their final form prior to publication [@prerelease].

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- NHS England Statistical Work Areas Division.
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- Via the NHS England website contact page or at the dmail: england.rtt@nhs.net
3. *Is there an erratum? If so, please provide a link or other access point.*
	- Errata are published with updates when necessary. 
	- An example erratum for October, November, December 2022 is on the 2022-2023 RTT page [@erratum]. 
	- All updates to guidance is tracked on the RTT-Guidance page [@guidance].

4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- Revisions are published periodically (usually every six months) in line with NHS England Analytical Service team’s revisions policy.
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- No retention limits are specified.
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Historical data until 2010 accessible on the site. 
	- Archived data until 2007 is available on the national archive site [@archive].   
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- No external contributions are accepted.
8. *Any other comments?*
	- Not applicable

\newpage

# References

### 2024-11-26 paper/paper.qmd
Wrote datasheet for RTT data.

@@
 ### Relationship between Wait Times and Type 1 Major Attendance
-
 The weak positive association between attendance rates and wait times, though not significant, aligns with broader concerns about rising patient loads straining healthcare systems.
+The relationship is likely weak due to the significant deviation from pre-Covid trend in 2020, with a sudden drop by 25% from 2019 levels.
+This drop was recovered in 2021, but the impact on the linear model constructed by a limited 8 observation data set cannot be ignored.
 As attendance increases, hospitals may prioritize emergency cases, diverting resources from elective and non-urgent procedures, potentially exacerbating delays.
@@
 To address the limitations of this study, future research should prioritize collecting monthly data for the years 2015 to 2019 to explore more granular trends before the anomalies introduced by COVID-19.
+@sec-adapted shows the potential for this further investigation with a model consisting of the yearly data from 2015 to 2019 exhibiting a much stronger $R^2$ value of 0.875.
+This model is small (5 observations), thus a more extensive model involving the monthly data would be beneficial.
 Analyzing these pre-pandemic patterns could provide a clearer understanding of the role that metrics such as beds per 1,000, physician density, and attendance rates played during a relatively stable period.
@@
 ```
-### Adapted Model 
+### Adapted Model {#sec-adapted}
 ```{r}

### 2024-11-26 paper/paper.qmd
Updated final paper abstract and introduction.

@@
 date-format: long
-abstract: "Hospital wait times in the United Kingdom are a critical measure of healthcare efficiency, reflecting the balance between resources and demand.
-This study examines the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for seven key medical procedures from 2015 to 2019.
-Using regression modeling, we find that reduced hospital beds per capita significantly prolong wait times,and while higher physician availability helps, it does little to offset the impact of declining bed availability.
-Additionally, Type 1 major care attendance rates correlate with increased delays, further indicating systemic bottlenecks.
-These findings highlight the urgent need for strategic resource allocation and demand management to address delays and improve patient outcomes."
+abstract: "Hospital wait times in the United Kingdom are a critical indicator of healthcare efficiency, reflecting the interplay between resources and patient demand.
+This study investigates how hospital infrastructure, workforce availability, and patient attendance rates influence wait times for fifteen key medical procedures between 2015 and 2022.
+Regression modeling reveals no statistically significant predictors, with hospital bed availability, physician density, and attendance rates showing weak associations with wait times.
+These findings suggest that systemic inefficiencies and external disruptions, such as the COVID-19 pandemic, play a more significant role than individual resource metrics.
+This study underscores the need for a deeper understanding of healthcare delays and highlights the importance of data-driven strategies to improve resource allocation and patient outcomes."
 format: pdf
@@
 # Introduction
-
 Timely access to healthcare is fundamental to achieving optimal health outcomes.
@@
 
-The findings reveal that declining hospital bed availability has a pronounced negative impact on wait times, while higher physician density and attendance rates are associated with additional delays, potentially due to demand outstripping capacity.
-These results emphasize the need for a balanced approach to resource allocation, combining infrastructure investment with demand management strategies.
+The findings reveal that none of the predictors have a statistically significant association with wait times, highlighting the complexity of the relationships between healthcare resources, workforce availability, and patient demand.
+Declining hospital bed availability, despite being hypothesized to have a strong effect, shows only a weak positive relationship with wait times, suggesting that other systemic factors, such as bed utilization efficiency and discharge planning, may overshadow its direct impact.
+Similarly, higher physician density is associated with increased delays, potentially reflecting inefficiencies in resource allocation or increased demand generated by greater access.
+Attendance rates also show a weak positive association, but the substantial deviation from pre-COVID trends, particularly the sharp drop in 2020 and rebound in 2021, likely undermines the robustness of the findings.
 
 Understanding the drivers of wait times is vital for improving healthcare access and efficiency [@king].
-By identifying the factors contributing to delays, this paper offers actionable insights for policymakers and healthcare administrators seeking to enhance patient outcomes and system resilience.
+By highlighting the multifaceted nature of healthcare delays and the limitations of current models, this paper underscores the need for more granular data and advanced methodologies to uncover the true determinants of wait times.
 The remainder of this paper is structured as follows: @sec-data details the data and methodology, @sec-model presents the model used for analysis of the data, @sec-results presents the key results from the model, and @sec-discussion concludes with final analysis of the results and recommendations for future research.

### 2024-11-28 paper/paper.qmd
Added literature, final references, and completed model.

@@
 ---
-title: "Decoding UK Hospital Wait Times: Infrastructure, Workforce, and Demand"
-subtitle: "How Systemic Pressures Shape Delays in Medical Care"
+title: "Decoding UK Hospital Wait Times: Systemic Challenges Beyond Resources"
+subtitle: "How Infrastructure, Workforce, and Demand do not Explain Delays in Medical Care"
 author: 
@@
 In the United Kingdom, the National Health Service (NHS) faces mounting pressure to manage increasing patient demands amidst resource constraints.
-Hospital wait times, a key performance indicator, have become a focal point for policymakers and stakeholders alike.
+Hospital wait times, a key performance indicator, have become a focal point for policymakers and stakeholders alike [@health].
 Extended wait times not only compromise patient care but also signal underlying inefficiencies in healthcare delivery systems.
@@
 This paper investigates how hospital infrastructure, workforce availability, and patient demand influence wait times for medical procedures across the UK.
-Specifically, it examines the roles of hospital beds per capita, physicians per 1,000 inhabitants, and Type 1 attendance rates in shaping delays for procedures like knee replacements, coronary artery bypass grafts, and cataract surgeries.
-While previous studies have explored individual contributors to wait times, this analysis integrates multiple predictors to provide a holistic view of systemic pressures.
+Specifically, it examines the roles of hospital beds per capita, physicians per 1,000 inhabitants, and Type 1 major care attendance rates in shaping delays from 2015 to 2022 for fifteen major fields of medicine presented in @tbl-treatments.
+The NHS consistently met the 18-week standard from 2008 to 2015, but has not since, thus the years 2015 on wards were selected for investigation.
+NHS waiting times continue to sharply increase making this a national healthcare crisis in the UK [@lancet].
 
-The estimand in this study is the average percentage change in wait times for seven key medical procedures: knee replacements, hip replacements, prostatectomy, hysterectomy, coronary artery bypass grafts, transluminal coronary angioplasty, and cataract surgeries in the UK.
+The estimand in this study is the average percentage change from 2015 in wait times for fifteen key fields of medicine @tbl-treatments.
 The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
@@
 
-In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR], using the haven and `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al.
-2021).
-All figures in the report are generated using `ggplot2` [@citeGG] and `patchwork` [@patchwork].
-We run the model in R [@citeR] using the `modelsummary` package of [@modelsummary].
+In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR] and use the `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al.
+2021) packages.
+All figures in the report are generated using `ggplot2` [@citeGG] with `patchwork` [@patchwork], and `gridExtra` [@gridExtra].
 
-This report integrates data from two key sources: the NHS and the OECD. These datasets provide critical information to analyze trends in healthcare demand and resources in the United Kingdom.
+This report integrates data from two key sources: the NHS and the Organisation for Economic Co-operation and Development (OECD).
+These datasets provide critical information to analyze trends in healthcare demand and resources in the United Kingdom.
 
 The NHS Referral to Treatment (RTT) waiting times dataset [@wait] is the primary source for waiting time statistics.
-These data measure the median time (in weeks) between referral by a general practitioner (GP) and the start of treatment for elective care, capturing delays across 15 key medical procedures such as general surgery, urology, and trauma & orthopedics.
+These data measure the median time (in weeks) between referral by a general practitioner (GP) and the start of treatment for elective care, capturing delays across 15 key medical procedures @tbl-treatments.
 This dataset offers procedure-specific insights that allow for cross-procedure and temporal comparisons.
+The data are aggregated and published monthly.
 
@@
 These statistics are collected at the provider organization level from NHS Trusts, Foundation Trusts, and independent sector organizations.
-The data are aggregated from monthly submissions, a system adopted following Sir Bruce Keogh’s review of waiting time standards, and include attendance and admission figures.
+The data are aggregated from monthly submissions.
 
 Additional datasets are sourced from the OECD Health Statistics 2024 database [@health], which provides standardized, internationally comparable data on healthcare systems.
-For the United Kingdom, the OECD compiles data on hospital resources hospital beds and physician counts from NHS Digital and Public Health Scotland.
-The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments, making it a reliable complement to NHS data for the analysis.
+For the United Kingdom, the OECD compiles data on hospital resources hospital beds and physician counts from NHS Digital and Public Health Scotland [@scotland].
+The OECD ensures consistency and comparability of this data through rigorous quality checks and methodological adjustments.
 
@@
 The data are collected and reported monthly through NHS administrative systems, with automated validation checks ensuring accuracy and consistency.
-Similar datasets include WHO and Eurostat health statistics offer broader regional comparisons but lack the granularity needed for a UK-focused study.
-[@https://turasdata.nes.nhs.scot/about-our-data-and-reports/methods/nhsscotland-workforce-data-warehouse/]
+The population data is taken from population estimates as those registered with the NHS.
 
@@
 It tracks the availability of inpatient beds.
-Annual averages of beds available overnight in public hospitals.
+It provides annual averages of beds available overnight in public hospitals.
 Includes acute care and psychiatric beds but excludes private sector facilities for consistency.
 Data are for financial years and represent publicly funded healthcare infrastructure.
-No significant methodological breaks are noted for this variable.
+The number of beds data is gathered through automated administrative streams from NHS hospitals and care facilities.
+The population data is taken from population estimates as those registered with the NHS.
 
@@
 **Overall**: All datasets rely on administrative data validated by their respective agencies, minimizing reporting errors.
-While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2019.
-Similar datasets from the World Health Organization (WHO) could offer supplementary insights but were not utilized due to a lack of UK-specific detail.
+While historical methodological changes (e.g., hospital reporting standards) necessitated adjustments, these do not affect the study's time period from 2015 to 2022.
+Similar datasets from the World Health Organization (WHO) and Eurostat health statistics could offer supplementary insights but were not utilized due to a lack of UK-specific detail.
 
 ## Methodology {#sec-methodology}
-The data cleaning method is described in @sec-cleaning.
+The data cleaning method involved many steps and is described in @sec-cleaning.
 All data cleaning was done using `arrow` [@arrow], `tidyverse` [@tidyverse], `readxl` [@readxl], and `stringr` [@stringr].
@@
 # Display the first part of the table
-kable(head(part1), 
-      caption = "Part 1: Hospital statistics")
+kable(part1, 
+      caption = "NHS Hospital Statistics for 2015-2022")
 ```
 
-@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of 10 variables and 8 observations spanning the years 2015 to 2022.
-The variables include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
-Additional variables capture wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery.
-In @tbl-data we see the average percentage change in wait times from 2015 levels in the `Total` column.
+@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of the 3 predictor variables and estimand and 8 observations spanning the years 2015 to 2022.
+The predictors include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
+The outcome variable `Total` captures wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery.
 All metrics are standardized based on population or healthcare utilization data for each year.
+The full cleaned dataset is provided in @ tbl-data-appendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.
 
 ## Outcome variables
-
-@fig-data depicts the percentage increase in wait times for various medical procedures compared to 2015.
-By 2019, most procedures show a notable increase in wait times, with angioplasty and prostatectomy exhibiting the largest rises.
-While some procedures, such as knee and hip surgeries, show consistent increases, others, like prostatectomy, initially decline before rising sharply.
-The steady increase across most procedures highlights a systemic trend of growing delays in accessing medical care, potentially signaling resource limitations or increased demand in healthcare services.
-Overall, wait times have risen by over 20% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.
 
@@
 These trends underscore the urgent need for targeted healthcare interventions, including expanded capacity, streamlined workflows, and policies to address the post-pandemic backlog and increasing demand.
+Overall, wait times have risen close to 30% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.
+
+\newpage
+
 ## Predictor variables
@@
 
+\newpage
+
 ### Physicians per 1,000
@@
 While this trend highlights progress in enhancing workforce availability, further analysis is needed to assess whether these increases adequately meet rising healthcare demands and address regional disparities in physician distribution.
+
+\newpage
 
@@
 
+\newpage
+
 # Model {#sec-model}
@@
 
+The model was run in `R` [@citeR] using the `modelsummary` package [@modelsummary].
+
 ## Model set-up
@@
 X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
-\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \
+\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \\
 \beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
@@
 
-Each predictor variable is expected to contribute uniquely:
+### Model Assumptions {#sec-model-assumptions}
+The linear regression model assumes:
 
-- Beds per 1,000 inhabitants (X_1): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
-- Physicians per 1,000 inhabitants (X_2): Greater physician availability is hypothesized to decrease wait times.
-- Attendance Rates (X_3): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.
-
-We run the model in R [@citeR] using the `modelsummary` package of `rsaratnam` [@rsaratnam].
+* Linearity: The relationship between the predictors and response variable is linear.
+* Independence: Observations are independent of each other.
+* Homoscedasticity: The variance of the error term (\epsilon) is constant across all levels of the predictors.
+* Normality: The residuals of the model are approximately normally distributed.
 
@@
 This analysis will help identify key leverage points for optimizing resource allocation and reducing wait times across various medical procedures.
+
+### Hypotheses
+Each predictor variable is expected to contribute uniquely:
+
+* Beds per 1,000 inhabitants (X_1): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
+* Physicians per 1,000 inhabitants (X_2): Greater physician availability is hypothesized to decrease wait times.
+* Attendance Rates (X_3): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.
+
+### Evaluation and Validation
+Evaluation and Validation
+To ensure the robustness of our findings, the model was validated through the following steps:
+
+1. Residual Diagnostics: We assessed residual plots for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
+2. Significance Testing: We conducted t-tests to assess whether each coefficient (\beta_0, \beta_1, \beta_2, \beta_3) is statistically significant at the 0.05 level @sec-results.
+3. Goodness of Fit: The R-squared statistic was calculated to measure how well the model explains the variability in wait times @sec-results.
+4.
+Alternative Models: a variant containing data only from 2015 to 2019 was tested to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit @sec-adapted.
+In both models, the p-values of the coefficients were insignificant so the model containing the larger data set (original model from 2015-2022) was used for final analysis.
 
@@
 ```
-The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (Pct_Wait_Avg) compared to three key predictors: attendance, physicians_per_1000, and beds_per_1000.
+The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (`Total`) compared to the three predictors: attendance, physicians_per_1000, and beds_per_1000.
 The red lines depict the predicted relationships, while the blue points represent the observed data values.
-The attendance predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend.
-Physicians_per_1000 also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance.
+The physicians_per_1000 predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend.
+Attendance also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance.
 In contrast, beds_per_1000 shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend.
-These plots suggest that attendance is the most influential predictor of wait times among the three variables, while beds_per_1000 has the least impact.
-
-@fig-modeldetails presents the residual plots.
-The residual plots indicate that the linear model generally fits the data well, with residuals scattered randomly around zero for all predictors, suggesting no major violations of linearity or independence assumptions.
-However, in each residual plot there is a violation of constant variance where there are noticeable outliers and potential influential points, particularly in the plots for attendance and physicians per 1,000, which may affect the model's robustness.
-The fitted vs.
-actual values plot shows a good overall alignment with most points near the predicted line, but two significant deviations highlight discrepancies that could stem from these outliers.
-Given the small sample size, these findings should be interpreted with caution.
+For each plot of the predictor vs outcome variable there seems to be at least two data points which deviate significantly from the remaining points.
+These were hypothesized to be due to the outliers from trend witnessed in @sec-data stemming from the Covid-19 pandemic years.
+@sec-adapted developed and analysed the significance of an adapted model for years 2015-2019 before the pandemic to avoid such outlier effects in capturing the trend.
+The overall model significance was much greater and linear assumption violations in the residual plots were non-existent, however individual coefficients were still non-significant and the dataset on which this model was constructed was even smaller with 5 observations.
 
@@
 ## Overview
-This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for key medical procedures in the United Kingdom from 2015 to 2022.
+This study investigates the relationships between healthcare infrastructure, workforce availability, and patient attendance rates with average wait times for 15 medical sectors in the United Kingdom from 2015 to 2022.
 Our analysis employs a linear regression model but finds no statistically significant relationships between the predictors and wait times.
@@
 
+### Treatments {#sec-treatments}
+
+```{r}
+#| echo: false
+#| warning: false
+#| error: false
+#| label: tbl-data-appendix
+#| tbl-cap: "A summary table of the treatments"
+
+# Import the cleaned healthcare dataset
+cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+
+# Extract column names starting from the 5th column to the second last column
+treatment_names <- colnames(cleaned_data)[5:(ncol(cleaned_data) - 1)]
+
+# Display the treatment names in a table
+library(knitr)
+kable(
+  data.frame("Treatments" = treatment_names),
+  caption = "List of Medical Treatment Areas"
+)
+```
+
+### Full Cleaned Dataset
 ```{r}
@@
 #| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
-# Import the cleaned healthcare dataset
-cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
 
@@
 # Display the first part of the table
-kable(head(part1), 
+kable(part1, 
       caption = "Part 1: Hospital statistics")
-kable(head(part2), 
+kable(part2, 
       caption = "Part 2: Hospital statistics")
-kable(head(part3), 
+kable(part3, 
       caption = "Part 3: Hospital statistics")
-kable(head(part4), 
+kable(part4, 
       caption = "Part 4: Hospital statistics")
+
 ```
@@
 ## Model Details {#sec-model-details}
-### Original Model Residuals
+### Original Model Residuals {#sec-residuals}
 ```{r}
 #| echo: false
-#| output: asis
+#| output: false
 #| eval: true
@@
 #| tbl-cap: "Residual plots of each predictor."
-#| # Load necessary libraries
-library(ggplot2)
 
@@
 ```
+@fig-modeldetails presents the residual plots which indicate that the linear model generally fits the data well, with residuals scattered randomly around zero for all predictors, suggesting no major violations of linearity or independence assumptions.
+However, in each residual plot there is a violation of constant variance where there are noticeable outliers and potential influential points, particularly in the plots for attendance and physicians per 1,000, which may affect the model's robustness.
+The fitted vs.
+actual values plot shows a good overall alignment with most points near the predicted line, but two significant deviations highlight discrepancies that could stem from these outliers.
+Given the small sample size, these findings should be interpreted with caution.
 ### Adapted Model {#sec-adapted}
@@
 #| eval: true
-#| label: tbl-modeladj
-#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors"
-
-#### Workspace setup ####
-library(tidyverse)
+#| label: tbl-adj
+#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"
 
@@
 # Load the cleaned data
-analysis_data <- read_csv("../data/02-analysis_data/final_healthcare_data.csv")
+analysis_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
 
 #### Filter data ####
-# Exclude data from 2020 to 2022
+# Include only data from 2015 to 2019
 filtered_data <- analysis_data %>% 
-  filter(Year < 2020)
+  filter(Year >= 2015 & Year <= 2019)
 
@@
 )
+```
+The linear regression model, summarized in @tbl-adj, examines the impact of healthcare system factors on wait times from 2015–2019.
+Hospital beds per 1,000 inhabitants (\beta_1 =−176.99) and physicians per 1,000 (\beta_2 =−78.19) both show negative associations with wait times, suggesting increased resources reduce delays, though neither effect is statistically significant (p >0.05).
+Attendance rates (\beta_3=−0.66) exhibit a minimal effect.
+The model explains 97.8% of the variability in wait times ($R^2$=0.978), though the adjusted $R^2$ of 0.911 reflects possible overfitting due to the small sample size (n = 5).
+While the model fits well, the lack of significant predictors highlights the need for further data to better capture drivers of wait times.
+The small RMSE of 0.58 indicates the linear trend is never far off from the actual wait times.
+
+```{r}
+#| echo: false
+#| output: false
+#| eval: true
+#| label: fig-residadj
+#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"
+# Extract residuals and fitted values from the model
+filtered_data <- filtered_data %>%
+  mutate(
+    residuals = residuals(modela),
+    fitted = fitted.values(modela)
+  )
+
+#### Diagnostic Plots ####
+
+# Plot b: Residuals vs Predictor 1 (attendance)
+plot_b <- ggplot(filtered_data, aes(x = attendance, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Attendance",
+    y = "Residuals",
+    x = "Attendance"
+  )
+
+# Plot c: Residuals vs Predictor 2 (physicians_per_1000)
+plot_c <- ggplot(filtered_data, aes(x = physicians_per_1000, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Physicians per 1,000",
+    y = "Residuals",
+    x = "Physicians per 1,000"
+  )
+
+# Plot d: Fitted values vs Actual values (Total)
+plot_e <- ggplot(filtered_data, aes(x = Total, y = fitted)) +
+  geom_point(color = "blue") +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
+  theme_classic() +
+  labs(
+    title = "Fitted vs Actual Values",
+    y = "Fitted Values",
+    x = "Actual Values"
+  )
+
+# Plot e: Residuals vs Predictor 3 (beds_per_1000)
+plot_d <- ggplot(filtered_data, aes(x = beds_per_1000, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Beds per 1,000",
+    y = "Residuals",
+    x = "Beds per 1,000"
+  )
+
+# Combine all plots into a grid
+combined_plot <- (plot_b | plot_c) / (plot_d | plot_e) +
+  plot_annotation(title = "Diagnostic Plots for Linear Model (2015-2019)")
+
+# Print the combined plot
+print(combined_plot)
+```
+@fig-resid-adj residuals show no clear patterns against attendance, physicians per 1,000, or beds per 1,000, supporting the assumption of linearity.
+There are no violations against constant variance, unlike in @fig-resid.
+The fitted values reasonably align with the actual values.
+However, the limited data points (5 observations) reduce the robustness of these conclusions.
+```{r}
+#| echo: false
+#| output: false
+#| eval: true
+#| label: fig-adjmodelresults
+#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)."
+
+# Function to create a projection plot for a specific predictor
+create_projection_plot <- function(predictor_name, data, model) {
+  # Create a grid of values for the predictor of interest
+  grid <- seq(min(data[[predictor_name]], na.rm = TRUE), 
+              max(data[[predictor_name]], na.rm = TRUE), length.out = 100)
+
+  # Create a new data frame for prediction
+  pred_data <- data.frame(grid)
+  colnames(pred_data) <- predictor_name
+
+  # Fix all other predictors at their mean values
+  other_predictors <- setdiff(names(model$coefficients)[-1], predictor_name)
+  for (var in other_predictors) {
+    pred_data[[var]] <- mean(data[[var]], na.rm = TRUE)
+  }
+
+  # Predict using the model
+  pred_data$Predicted <- predict(model, newdata = pred_data)
+
+  # Create the plot
+  ggplot(data, aes_string(x = predictor_name, y = "Total")) +
+    geom_point(color = "blue", size = 2, alpha = 0.7) +
+    geom_line(data = pred_data, aes_string(x = predictor_name, y = "Predicted"), 
+              color = "red", size = 1) +
+    labs(
+      x = predictor_name,
+      y = "Wait Times (%)"
+    ) +
+    theme_minimal() +
+    theme(aspect.ratio = 0.8)
+}
+
+# Generate individual plots using filtered data (2015-2019)
+p1 <- create_projection_plot("attendance", filtered_data, modela)
+p2 <- create_projection_plot("physicians_per_1000", filtered_data, modela)
+p3 <- create_projection_plot("beds_per_1000", filtered_data, modela)
+
+# Combine the plots into one figure
+combined_plot <- (p1 | p2 | p3) +
+  plot_annotation(
+    title = "Projections of Linear Model (2015-2019) onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
+    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
+  )
+
+# Display the combined plot
+print(combined_plot)
 
 ```
+Overall, the linear model projections in @fig-adjmodelresults effectively captures general trends, particularly for hospital bed availability.
+However, the variability around the red lines, particularly for attendance and physicians per 1,000, indicates potential model limitations or missing interactions between predictors.
+The small sample size (n = 5) further limits the reliability of these insights, necessitating caution when generalizing these findings.
+Additional data or non-linear modeling may better address these discrepancies and provide a more comprehensive understanding of the relationships between these predictors and wait times.
 

### 2024-11-28 scripts/02-clean_data.R

@@ -121,15 +121,6 @@ yearly_averages <- all_data %>%
 wide_yearly_averages <- yearly_averages %>%
   pivot_wider(names_from = treatment_function, values_from = Average_Waiting_Time)
 
-wide_yearly_averages <- wide_yearly_averages %>%
-  mutate(
-    ENT = case_when(
-      Year == 2019 ~ 14.9502,  # Set the value for 2019
-      Year == 2020 ~ 15.216,   # Set the value for 2020
-      TRUE ~ ENT              # Retain existing values for other years
-    )
-  )
-
 # Calculate percent change from 2015 levels with robust handling of NA values
 wide_yearly_averages_percent_change <- wide_yearly_averages %>%
   mutate(across(
@@ -237,9 +228,11 @@ final_data <- physician_data |>
     Trauma,
     Urology,
     Total
+  ) 
   )
 
 
 #### Save final dataset ####
 write_parquet(final_data, "./data/02-analysis_data/final_healthcare_data.parquet")
 write.csv(final_data, "./data/02-analysis_data/final_healthcare_data.csv")
+

### 2024-11-28 paper/paper.qmd
Completed paper with appendix.

@@
 
-In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR] and use the `tidyverse` [@citeTidy], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al.
+In this paper, the analysis will be carried out using the statistical programming language `R` [@citeR] and use the `tidyverse` [@tidyverse], `devtools` (Wickham, Hester, and Chang 2020) and `dplyr` (Wickham et al.
 2021) packages.
@@
 
+\newpage
+
 ## Methodology {#sec-methodology}
@@
 Below is a snapshot of the first five rows of the key data columns which will be used for this paper.
-@tbl-data-appendix provides the entire cleaned dataset in table format.
+@tbl-dataappendix provides the data including individual treatment area percentage change in waiting times from 2015 in table format.
 
@@
 #| label: tbl-data
-#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
+#| tbl-cap: "NHS Hospital Statistics from the cleaned healthcare data for 2015-2022"
 # Import the cleaned healthcare dataset
@@
 
+format_numeric <- function(df) {
+  df %>%
+    mutate(across(where(is.numeric), ~ round(., 2))) # Round numeric columns to 2 decimal places
+}
+
 # Split the table into smaller parts
-part1 <- cleaned_data[, c(1:4, which(colnames(cleaned_data) == "Total"))]
+part1 <- cleaned_data[, c(1:4, which(colnames(cleaned_data) == "Total"))] %>% format_numeric
 
 # Display the first part of the table
-kable(part1, 
-      caption = "NHS Hospital Statistics for 2015-2022")
+kable(part1)
 ```
@@
 All metrics are standardized based on population or healthcare utilization data for each year.
-The full cleaned dataset is provided in @ tbl-data-appendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.
+The dataset including individual treatment waiting times percetentage change from 2015 is provided in @tbl-dataappendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.
+
+\newpage
 
@@
 # Plot for the "Total" column
-ggplot(long_wait_pct_data %>% filter(Treatment == "Total"), 
-       aes(x = Year, y = Pct_Change, group = Treatment)) +
+ggplot(cleaned_data , 
+       aes(x = Year, y = Total)) +
   geom_line(size = 1, color = "black") +
@@
     legend.position = "none",
-    text = element_text(size = 14)
+    text = element_text(size = 11)
   )
@@
 ggplot(cleaned_data) +
-  geom_line(aes(x = Year, y = attendance, color = "A&E Attendances"), size = 1) +
+  geom_line(aes(x = Year, y = attendance), size = 1) +
   scale_color_manual(values = c("A&E Attendances" = "blue")) +
@@
   theme_minimal() +
-  theme(legend.position = "bottom")
+  theme(legend.position = "bottom", text = element_text(size = 11))
 ```
@@
 
-\begin{align} 
-Y_{ij} & = \text{The average percentage change in wait time for the 15 specific medical procedures in year i}. \\
-X_{1i}, X_{2i}, X_{3i} & = \text{The independent variables of physicians per 1000 people, beds per 1000 people, and Type 1 attendance for procedure } j \text{ in year } i. \\
-\beta_0 & = \text{The intercept or constant term, representing the expected wait time when all predictors are zero.} \\
-\beta_1, \beta_2, \beta_3 & = \text{The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.} \\
-\epsilon_{i} & = \text{The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.}
-\end{align}
+* $Y_{i}$: The average percentage change in wait time for the 15 specific medical procedures in year $i$. 
+* $X_{1i}, X_{2i}, X_{3i}$: The independent variables representing physicians per 1,000 people, beds per 1,000 people, and Type 1 attendance, respectively, for procedure $j$ in year $i$. 
+* $\beta_0$: The constant term, representing the expected wait time when all predictors are zero.
+* $\beta_1, \beta_2, \beta_3$: The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.
+* $\epsilon_{i}$: The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.
 
-The goal of the linear regression model is to estimate the values of \beta_1, \beta_2, and \beta_3 such that the model fits the data well and to predict the expected value of the average total waiting time for procedures for different values of physician rates, hospital bed rates, and type 1 major care attendance.
+The goal of the linear regression model is to estimate the values of $\beta_1$, $\beta_2$, and $\beta_3$ such that the model fits the data well and to predict the expected value of the average total waiting time for procedures for different values of physician rates, hospital bed rates, and type 1 major care attendance.
 The statistical significance of 1 can be assessed using a t-test, which tests whether the estimated coeﬀicient is significantly different from zero.
@@
 * Independence: Observations are independent of each other.
-* Homoscedasticity: The variance of the error term (\epsilon) is constant across all levels of the predictors.
+* Homoscedasticity: The variance of the error term ($\epsilon$) is constant across all levels of the predictors.
 * Normality: The residuals of the model are approximately normally distributed.
@@
 The results from this model can provide actionable insights to inform policies aimed at optimizing resource allocation, improving healthcare infrastructure, and managing patient flow to reduce wait times.
-
-We expect to confirm that reduced healthcare resources (fewer hospital beds, lower physician availability), and increased demand (more tpe 1 attendances) are associated with longer wait times, while better resource availability can mitigate delays.
 This analysis will help identify key leverage points for optimizing resource allocation and reducing wait times across various medical procedures.
@@
 
-* Beds per 1,000 inhabitants (X_1): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
-* Physicians per 1,000 inhabitants (X_2): Greater physician availability is hypothesized to decrease wait times.
-* Attendance Rates (X_3): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.
+* Beds per 1,000 inhabitants ($X_1$): A lower bed-to-population ratio may increase wait times by limiting inpatient care capacity.
+* Physicians per 1,000 inhabitants ($X_2$): Greater physician availability is hypothesized to decrease wait times.
+* Attendance Rates ($X_3$): Increased attendance could indicate growing demand on the healthcare system, potentially leading to longer wait times.
 
@@
 1. Residual Diagnostics: We assessed residual plots for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
-2. Significance Testing: We conducted t-tests to assess whether each coefficient (\beta_0, \beta_1, \beta_2, \beta_3) is statistically significant at the 0.05 level @sec-results.
+2. Significance Testing: We conducted t-tests to assess whether each coefficient ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$) is statistically significant at the 0.05 level @sec-results.
 3. Goodness of Fit: The R-squared statistic was calculated to measure how well the model explains the variability in wait times @sec-results.
@@
 
+\newpage
+
 # Results {#sec-results}
@@
 #| echo: false
-#| output: asis
+#| warning: false
 #| eval: true
@@
 The linear regression model, presented in @tbl-modelresults, explores the relationships between healthcare system factors and the percentage change in wait times, though none of the variables demonstrate statistically significant effects, as indicated by p-values greater than 0.05.
-The number of hospital beds per 1,000 inhabitants shows a small positive effect (\beta=4.71), suggesting a minimal increase in wait times with higher bed availability, though this effect lacks statistical significance.
-Physician availability per 1,000 (\beta=31.83) indicates a larger positive association, yet it is also not statistically significant.
-Attendance rates (\beta =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times.
+The number of hospital beds per 1,000 inhabitants shows a small positive effect ($\beta_1$=4.71), suggesting a minimal increase in wait times with higher bed availability, though this effect lacks statistical significance.
+Physician availability per 1,000 ($\beta_2$=31.83) indicates a larger positive association, yet it is also not statistically significant.
+Attendance rates ($\beta_3$ =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times.
 The model achieves an $R^2$ value of 0.599, indicating that approximately 60% of the variability in wait times is explained by the model, though the adjusted $R^2$ value of 0.299 reflects limited predictive power.
@@
 
+\newpage
+
 ```{r}
 #| echo: false
-#| output: asis
+#| warning: false
+#| error: false
 #| eval: true
@@
     caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
+  ) &
+  theme(
+    plot.title = element_text(size = 10)  # Adjust the title font size here
   )
@@
 For each plot of the predictor vs outcome variable there seems to be at least two data points which deviate significantly from the remaining points.
-These were hypothesized to be due to the outliers from trend witnessed in @sec-data stemming from the Covid-19 pandemic years.
+These were hypothesized to be due to the outliers from trend witnessed in @fig-total stemming from the Covid-19 pandemic years.
 @sec-adapted developed and analysed the significance of an adapted model for years 2015-2019 before the pandemic to avoid such outlier effects in capturing the trend.
@@
 Additionally, the dataset is influenced by the effects of COVID-19, with wait times deviating significantly from the pre-pandemic trend.
-As illustrated in @fig-wait, wait times followed a consistent linear increase from 2015 to 2019, but 2020 deviated sharply below trend, while 2021 exhibited a sharp increase above trend.
+As illustrated in @fig-total, wait times followed a consistent linear increase from 2015 to 2019, but 2020 deviated sharply below trend, while 2021 exhibited a sharp increase above trend.
 These anomalies suggest inconsistent variance in the data and indicate that a linear model may not be appropriate for post-2019 data.
@@
 #| error: false
-#| label: tbl-data-appendix
+#| label: tbl-treatments
 #| tbl-cap: "A summary table of the treatments"
@@
 kable(
-  data.frame("Treatments" = treatment_names),
-  caption = "List of Medical Treatment Areas"
+  data.frame("Treatments" = treatment_names)
 )
 ```
-
-### Full Cleaned Dataset
-```{r}
-#| echo: false
-#| warning: false
-#| error: false
-#| label: tbl-data-appendix
-#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
-
-# Split the table into smaller parts
-part1 <- cleaned_data[, c(1:6)]
-part2 <- cleaned_data[, c(6:12)]
-part3 <- cleaned_data[, c(12:18)]
-part4 <- cleaned_data[, c(18:ncol(cleaned_data))]
-
-# Display the first part of the table
-kable(part1, 
-      caption = "Part 1: Hospital statistics")
-kable(part2, 
-      caption = "Part 2: Hospital statistics")
-kable(part3, 
-      caption = "Part 3: Hospital statistics")
-kable(part4, 
-      caption = "Part 4: Hospital statistics")
-
-```
-
-## Data Cleaning {#sec-cleaning}
+### Data Cleaning {#sec-cleaning}
 The data cleaning process undertaken in this project involved several key steps to ensure the dataset was accurate, standardized, and ready for analysis.
@@
 
-### Full Outcome Variable Plots
+### Individual Treatment Wait Times {#sec-cleaneddataset}
 ```{r}
@@
 #| error: false
-#| label: fig-data-group1
-#| tbl-cap: "Percentage Increase in Wait Times (Group 1)"
-#| fig-height: 8
-#| fig-width: 12
+#| label: tbl-dataappendix
+#| tbl-cap: "A summary table of the cleaned healthcare data displayed in parts"
 
-# Reshape the data to long format for plotting
-long_wait_pct_data <- cleaned_data |>
-  pivot_longer(
-    cols = -Year, # Include all columns except 'Year'
-    names_to = "Treatment",
-    values_to = "Pct_Change"
-  ) |>
-  filter(!Treatment %in% c("attendance", "physicians_per_1000", "beds_per_1000", "General_S"))
-
-# Split treatments into three groups and ensure "Total" is included in all
-treatments <- unique(long_wait_pct_data$Treatment)
-group_size <- ceiling((length(treatments) - 1) / 3)  # Exclude "Total" when dividing groups
-group1 <- c("Total", treatments[1:group_size])
-group2 <- c("Total", treatments[(group_size + 1):(2 * group_size)])
-group3 <- c("Total", treatments[(2 * group_size + 1):length(treatments)])
-
-# Define colors for treatments, ensuring "Total" is black
-generate_colors <- function(group) {
-  setNames(
-    c("black", rainbow(length(group) - 1)),
-    group
-  )
+format_numeric <- function(df) {
+  df %>%
+    mutate(across(where(is.numeric), ~ round(., 2))) # Round numeric columns to 2 decimal places
 }
 
-# Plot for the first group
-colors_group1 <- generate_colors(group1)
+# Split and format the table into smaller parts
+part1 <- cleaned_data[, c(1:6)] %>% format_numeric()
+part2 <- cleaned_data[, c(6:12)] %>% format_numeric()
+part3 <- cleaned_data[, c(12:18)] %>% format_numeric()
+part4 <- cleaned_data[, c(18:ncol(cleaned_data))] %>% format_numeric()
 
-plot1 <- long_wait_pct_data %>%
-  filter(Treatment %in% group1) %>%
-  ggplot(aes(x = Year, y = Pct_Change, color = Treatment, group = Treatment)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  scale_color_manual(
-    values = colors_group1,
-    guide = guide_legend(title = "Treatment")
-  ) +
-  labs(
-    title = "Percentage Increase in Wait Times (Group 1)",
-    x = "Year",
-    y = "Percentage Increase (%)"
-  ) +
-  theme_minimal() +
-  theme(legend.position = "bottom", text = element_text(size = 14))
-
-print(plot1)
+# Display the first part of the table
+kable(part1)
 
@@
 #| error: false
-#| label: fig-data-group2
-#| tbl-cap: "Percentage Increase in Wait Times (Group 2)"
-#| fig-height: 8
-#| fig-width: 12
+#| label: tbl-dataappendix2
+#| tbl-cap: "Part 2: Hospital statistics"
 
-# Plot for the first group
-colors_group2 <- generate_colors(group1)
+kable(part2)
 
-plot1 <- long_wait_pct_data %>%
-  filter(Treatment %in% group2) %>%
-  ggplot(aes(x = Year, y = Pct_Change, color = Treatment, group = Treatment)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  scale_color_manual(
-    values = colors_group1,
-    guide = guide_legend(title = "Treatment")
-  ) +
-  labs(
-    title = "Percentage Increase in Wait Times (Group 2)",
-    x = "Year",
-    y = "Percentage Increase (%)"
-  ) +
-  theme_minimal() +
-  theme(legend.position = "bottom", text = element_text(size = 14))
+```
 
-print(plot2)
+```{r}
+#| echo: false
+#| warning: false
+#| error: false
+#| label: tbl-dataappendix3
+#| tbl-cap: "Part 3: Hospital statistics"
+
+kable(part3)
 
@@
 #| error: false
-#| label: fig-data-group2
-#| tbl-cap: "Percentage Increase in Wait Times (Group 3)"
-#| fig-height: 8
-#| fig-width: 12
+#| label: tbl-dataappendix4
+#| tbl-cap: "Part 4: Hospital statistics"
 
-# Plot for the second group
-plot3 <- long_wait_pct_data %>%
-  filter(Treatment %in% group3) %>%
-  ggplot(aes(x = Year, y = Pct_Change, color = Treatment, group = Treatment)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  labs(
-    title = "Percentage Increase in Wait Times (Group 3)",
-    x = "Year",
-    y = "Percentage Increase (%)",
-    color = "Treatment"
-  ) +
-  theme_minimal() +
-  theme(legend.position = "bottom", text = element_text(size = 14))
+kable(part4)
 
-print(plot3)
 ```
+
+\newpage
+
 ## Model Details {#sec-model-details}
+
 ### Original Model Residuals {#sec-residuals}
@@
 #| echo: false
-#| output: false
+#| warning: false
+#| error: false
 #| eval: true
@@
     x = "Attendance"
-  )
+  ) +
+  theme(plot.title = element_text(size = 10))
 
@@
     x = "Physicians per 1,000"
-  )
+  ) +
+  theme(plot.title = element_text(size = 10))
 
@@
     x = "Actual Values"
-  )
+  ) +
+  theme(plot.title = element_text(size = 10))
 
@@
     x = "Beds per 1,000"
-  )
+  ) +
+  theme(plot.title = element_text(size = 10))
 
@@
 Given the small sample size, these findings should be interpreted with caution.
+
+\newpage
+
 ### Adapted Model {#sec-adapted}
+
 ```{r}
@@
 #| output: asis
+#| warning: false
 #| eval: true
@@
 The linear regression model, summarized in @tbl-adj, examines the impact of healthcare system factors on wait times from 2015–2019.
-Hospital beds per 1,000 inhabitants (\beta_1 =−176.99) and physicians per 1,000 (\beta_2 =−78.19) both show negative associations with wait times, suggesting increased resources reduce delays, though neither effect is statistically significant (p >0.05).
-Attendance rates (\beta_3=−0.66) exhibit a minimal effect.
-The model explains 97.8% of the variability in wait times ($R^2$=0.978), though the adjusted $R^2$ of 0.911 reflects possible overfitting due to the small sample size (n = 5).
+Hospital beds per 1,000 inhabitants ($\beta_1 =−176.99$) and physicians per 1,000 ($\beta_2 =−78.19$) both show negative associations with wait times, suggesting increased resources reduce delays, though neither effect is statistically significant ($p >0.05$).
+Attendance rates ($\beta_3=−0.66$) exhibit a minimal effect.
+The model explains 97.8% of the variability in wait times ($R^2$=0.978), though the adjusted $R^2$ of 0.911 reflects possible overfitting due to the small sample size ($n = 5$).
 While the model fits well, the lack of significant predictors highlights the need for further data to better capture drivers of wait times.
@@
 
+\newpage
+
 ```{r}
 #| echo: false
-#| output: false
+#| warning: false
 #| eval: true
@@
 ```
-@fig-resid-adj residuals show no clear patterns against attendance, physicians per 1,000, or beds per 1,000, supporting the assumption of linearity.
-There are no violations against constant variance, unlike in @fig-resid.
+@fig-residadj residuals show no clear patterns against attendance, physicians per 1,000, or beds per 1,000, supporting the assumption of linearity.
+There are no violations against constant variance, unlike in @fig-modeldetails.
 The fitted values reasonably align with the actual values.
 However, the limited data points (5 observations) reduce the robustness of these conclusions.
+
+\newpage
+
 ```{r}
 #| echo: false
-#| output: false
+#| warning: false 
+#| error: false
 #| eval: true
 #| label: fig-adjmodelresults
-#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)."
+#| tbl-cap: "Projections of Linear Model onto Each Predictor (2015-2019)."
 

### 2024-12-03 paper/paper.qmd
Completed idealized methodology and LLM usage.

@@
 
-The estimand in this study is the average percentage change from 2015 in wait times for fifteen key fields of medicine @tbl-treatments.
+The estimand in this study is the is the median number of weeks spent waiting for treatment since referral averaged over fifteen key fields of medicine @tbl-treatments.
 The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
@@
 The predictors include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates.
-The outcome variable `Total` captures wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery.
+The outcome variable `Total` captures wait times (median number of weeks per year) averaged over key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery.
 All metrics are standardized based on population or healthcare utilization data for each year.
-The dataset including individual treatment waiting times percetentage change from 2015 is provided in @tbl-dataappendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.
+The dataset including individual treatment waiting times in weeks is provided in @tbl-dataappendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.
 
@@
 #| label: fig-total
-#| tbl-cap: "Percentage Increase in Total Wait Times Compared to 2015"
+#| tbl-cap: "Median Number of Weeks for Total Wait Times Compared to 2015"
 
@@
   labs(
-    title = "Percentage Change in Total Wait Times Compared to 2015",
+    title = "Median Number of Weeks for Total Wait Times Compared to 2015",
     x = "Year",
-    y = "Percentage Change (%)",
+    y = "Average Wait Time (Median # of Weeks)",
     color = NULL
@@
     legend.position = "none",
-    text = element_text(size = 11)
+    text = element_text(size = 10),
+    plot.title = element_text(size = 11),  
   )
@@
 #| label: fig-beds
-#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2019
+#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2022
 #| echo: false
@@
   geom_line(color = "blue") +
-  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
+  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2022", x = "Year", y = "Beds per 1 000 Inhabitants") +
   theme_minimal()
@@
 #| label: fig-physicians
-#| fig-cap: United Kingdom's Physician rate per 1000 Inhabitants from 2015 to 2019
+#| fig-cap: United Kingdom's Physician rate per 1000 Inhabitants from 2015 to 2022
 #| echo: false
@@
   geom_line(color = "blue") +
-  labs(title = "Physicians per 1 000 rates from 2015 to 2019", x = "Year", y = "Physicians per 1 000") +
+  labs(title = "Physicians per 1 000 rates from 2015 to 2022", x = "Year", y = "Physicians per 1 000") +
   theme_minimal()
@@
 #| label: fig-attendance-demand
-#| fig-cap: "Type 1 Major Care Attendance in the United Kingdom from 2015 to 2019"
+#| fig-cap: "Type 1 Major Care Attendance in the United Kingdom from 2015 to 2022"
 #| echo: false
@@
   labs(
-    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2019)",
+    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2022)",
     x = "Year",
@@
   theme_minimal() +
-  theme(legend.position = "bottom", text = element_text(size = 11))
+  theme(legend.position = "bottom", text = element_text(size = 10), plot.title = element_text(size = 11),  ) 
 ```
@@
 # Model {#sec-model}
-After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., physicians per 1,000 people, hospital beds per 1,000 people, and Type 1 attendance rates) and wait times for medical procedures.
+After conducting exploratory analysis on the dataset, linear trends were observed in each of the predictors and the outcome variable for years 2015 to 2019.
+Years 2020 to 2022 exhibited significant deviations from the previous years' linear trend, likely due to the Covid19 pandemic.
+Thus years 2015 to 2019 were thought to be the best to model.
+Linear relationships were observed between healthcare system factors ( physicians per 1,000 people, hospital beds per 1,000 people, and Type 1 attendance rates) and wait times for medical procedures.
 These variables indicate potential predictive power, suggesting a linear relationship.
-To further investigate and quantify how these factors influence wait times, we implemented a linear regression model.
+To further investigate and quantify how these factors influence wait times, we implemented a multivariate linear regression model.
 
@@
 
-* $Y_{i}$: The average percentage change in wait time for the 15 specific medical procedures in year $i$. 
+* $Y_{i}$: The median number of weeks averaged over the 15 specific medical procedures in year $i$. 
 * $X_{1i}, X_{2i}, X_{3i}$: The independent variables representing physicians per 1,000 people, beds per 1,000 people, and Type 1 attendance, respectively, for procedure $j$ in year $i$. 
@@
 
+These are checked in @sec-residuals.
+
 ### Model justification
@@
 This model framework allows us to evaluate the impact of key healthcare system factors on wait times.
-The selected predictors—beds_per_1000, physicians_per_1000, and attendance—are directly linked to resource availability and demand, making them relevant for understanding delays in medical care.
+The selected predictors $beds\_per\_1000$, $physicians\_per\_1000$, and $attendance$ are directly linked to resource availability and demand, making them relevant for understanding delays in medical care.
 The results from this model can provide actionable insights to inform policies aimed at optimizing resource allocation, improving healthcare infrastructure, and managing patient flow to reduce wait times.
@@
 ### Evaluation and Validation
-Evaluation and Validation
-To ensure the robustness of our findings, the model was validated through the following steps:
+To ensure the robustness of the findings, the model was validated through the following steps:
 
@@
 4.
-Alternative Models: a variant containing data only from 2015 to 2019 was tested to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit @sec-adapted.
-In both models, the p-values of the coefficients were insignificant so the model containing the larger data set (original model from 2015-2022) was used for final analysis.
+Alternative Models: a variant containing data from 2015 to 2022 was tested in @sec-adapted to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit.
+This model had a much smaller $R^2$ and adjusted $R^2$ values and a much larger RMSE.
+The observation size differences between the two models were not significant enough to account for the much poorer fit, thus the smaller model was kept for final analysis.
 
@@
 #| echo: false
+#| output: asis
 #| warning: false
 #| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors"
+#| label: tbl-adj
+#| tbl-cap: "Linear model explaining wait times (median number of weeks per year) based on healthcare system predictors (2015-2019)"
 
 # Load the saved model
-model <- readRDS("../models/wait_times_model.rds")
+modela <- readRDS("../models/wait_times_model_pre_2020.rds")
 
@@
 modelsummary::modelsummary(
-  list("Frequentist Linear Model" = model),
+  list("Multivariate Linear Model" = modela),
   statistic = c("p.value")
@@
 ```
-
-The linear regression model, presented in @tbl-modelresults, explores the relationships between healthcare system factors and the percentage change in wait times, though none of the variables demonstrate statistically significant effects, as indicated by p-values greater than 0.05.
-The number of hospital beds per 1,000 inhabitants shows a small positive effect ($\beta_1$=4.71), suggesting a minimal increase in wait times with higher bed availability, though this effect lacks statistical significance.
-Physician availability per 1,000 ($\beta_2$=31.83) indicates a larger positive association, yet it is also not statistically significant.
-Attendance rates ($\beta_3$ =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times.
-The model achieves an $R^2$ value of 0.599, indicating that approximately 60% of the variability in wait times is explained by the model, though the adjusted $R^2$ value of 0.299 reflects limited predictive power.
-With a sample size of n=8, the findings should be interpreted cautiously, as the lack of significance and low adjusted $R^2$ highlight potential limitations in the model's robustness and generalizability.
+The linear regression model, summarized in @tbl-adj, evaluates the impact of healthcare system factors on wait times from 2015–2019, revealing that hospital beds per 1,000 inhabitants ($\beta_1 = -16.826$) and physicians per 1,000 inhabitants ($\beta_2 = -7.196$) are negatively associated with wait times, indicating that increased resources may reduce delays however the findings are statistically insignificant ($p > 0.05$), highlighting the limitations imposed by the small sample size ($n = 5$).
+Attendance rates ($\beta_3 = -0.075$) exhibit a negligible effect, with a low p-value (0.714) further suggesting minimal impact on wait times when controlling for other factors.
+The model explains 96.6% of the variability in wait times ($R^2 = 0.966$), but the adjusted $R^2$ of 0.866 points to potential overfitting, as the inclusion of weak predictors may inflate the apparent fit.
+A small root mean squared error (RMSE) of approximately 0.07 indicates that the model's predictions closely align with observed values, yet this performance may not generalize beyond the dataset.
+Despite the excellent overall fit, the lack of significant predictors underscores the limitations of the analysis, with the small sample size undermining the precision of coefficient estimates and the robustness of conclusions.
+Furthermore, the variability in predictor effects suggests the potential presence of omitted variables or non-linear interactions.
 
@@
 #| eval: true
-#| label: fig-modelresults1
-#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining percentage increase in wait times based on healthcare system predictors."
-
-# Load the saved model
-model <- readRDS("../models/wait_times_model.rds")
-
-# Load the dataset used for the model
-data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+#| label: fig-adjmodelresults
+#| tbl-cap: "Projections of Linear Model onto Each Predictor (2015-2019)."
 
@@
       x = predictor_name,
-      y = "Wait Times (%)"
+      y = "Wait Times (# Weeks)"
     ) +
@@
 
-# Generate individual plots
-p1 <- create_projection_plot("attendance", data, model)
-p2 <- create_projection_plot("physicians_per_1000", data, model)
-p3 <- create_projection_plot("beds_per_1000", data, model)
+analysis_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+
+# Include only data from 2015 to 2019
+filtered_data <- analysis_data %>% 
+  filter(Year >= 2015 & Year <= 2019)
+
+# Generate individual plots using filtered data (2015-2019)
+p1 <- create_projection_plot("attendance", filtered_data, modela)
+p2 <- create_projection_plot("physicians_per_1000", filtered_data, modela)
+p3 <- create_projection_plot("beds_per_1000", filtered_data, modela)
 
@@
   plot_annotation(
-    title = "Projections of Linear Model onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
+    title = "Projections of Linear Model (2015-2019) onto \nEach Predictor vs Average Wait Times (Weeks)",
     caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
@@
   theme(
-    plot.title = element_text(size = 10)  # Adjust the title font size here
+    plot.title = element_text(size = 11),  
+    axis.title.y = element_text(size = 9)
   )
@@
 ```
-The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (`Total`) compared to the three predictors: attendance, physicians_per_1000, and beds_per_1000.
-The red lines depict the predicted relationships, while the blue points represent the observed data values.
-The physicians_per_1000 predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend.
-Attendance also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance.
-In contrast, beds_per_1000 shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend.
-For each plot of the predictor vs outcome variable there seems to be at least two data points which deviate significantly from the remaining points.
-These were hypothesized to be due to the outliers from trend witnessed in @fig-total stemming from the Covid-19 pandemic years.
-@sec-adapted developed and analysed the significance of an adapted model for years 2015-2019 before the pandemic to avoid such outlier effects in capturing the trend.
-The overall model significance was much greater and linear assumption violations in the residual plots were non-existent, however individual coefficients were still non-significant and the dataset on which this model was constructed was even smaller with 5 observations.
+The projection plots in @fig-adjmodelresults illustrate the linear model's predictions for the average wait timein relation to the three predictors: attendance, physicians per 1,000, and beds per 1,000.
+The red lines in each plot represent the predicted relationships, while the blue points reflect the observed data.
+For attendance, the model suggests a negative relationship with wait times, as indicated by the downward-sloping red line.
+However, the observed data points display considerable variability around the prediction, with notable deviations suggesting the presence of unmodeled interactions or confounding factors.
+The relationship for physicians per 1,000 is also negative, with a steep downward slope showing that higher physician density is associated with reduced wait times.
+While the data points align more closely with the model's predictions compared to attendance, some deviations, particularly at lower physician densities, highlight potential limitations in the model’s assumptions.
+The strongest and most consistent relationship is observed for beds per 1,000, where the steep downward-sloping red line indicates a significant association between increased bed availability and reduced wait times.
+The observed data points align closely with this trend, suggesting that this predictor has a robust and straightforward influence on wait times.
+
+Across all predictors, however, the small sample size (n = 5) significantly limits the reliability and generalizability of the findings.
+While the data for 2015–2019 avoids the confounding impact of the COVID-19 pandemic, the limited dataset reduces the ability to draw definitive conclusions, and individual coefficient estimates remain non-significant despite the improved model fit.
+The variability around the predictions, particularly for attendance and physicians per 1,000, highlights potential missing interactions or non-linear relationships that the linear model cannot capture.
+The model overall captures the relationship between the predictors and the outcome variable better than the model investigated in @sec-adapted.
 
@@
 
-The positive but statistically insignificant relationship between physician density and wait times challenges the expectation that greater workforce availability reduces delays.
+The negative but statistically insignificant relationship between physician density and wait times challenges the expectation that greater workforce availability reduces delays.
 This result may reflect systemic inefficiencies or increased demand driven by greater access to healthcare services.
@@
 ### Relationship between Wait Times and Type 1 Major Attendance
-The weak positive association between attendance rates and wait times, though not significant, aligns with broader concerns about rising patient loads straining healthcare systems.
+The weak negative association between attendance rates and wait times, though not significant, aligns with broader concerns about rising patient loads straining healthcare systems.
 The relationship is likely weak due to the significant deviation from pre-Covid trend in 2020, with a sudden drop by 25% from 2019 levels.
@@
 ## Limitations 
-The primary limitation of this study is the small sample size, which includes only eight observations spanning 2015 to 2022.
-This limits the generalizability of findings and weakens the statistical power needed to detect significant relationships.
-Additionally, the dataset is influenced by the effects of COVID-19, with wait times deviating significantly from the pre-pandemic trend.
+The primary limitation of this study is the small sample size ($n = 5$ for the focused 2015–2019 model), which weakens the statistical power and limits the generalizability of the findings.
+The dataset also includes anomalies introduced by the COVID-19 pandemic, with wait times deviating sharply from the pre-pandemic trend in 2020 and 2021.
+These anomalies introduce inconsistent variance in the data, further reducing the reliability of the adapted linear model in @sec-adapted, which assumes consistent patterns across the dataset.
 As illustrated in @fig-total, wait times followed a consistent linear increase from 2015 to 2019, but 2020 deviated sharply below trend, while 2021 exhibited a sharp increase above trend.
@@
 To address the limitations of this study, future research should prioritize collecting monthly data for the years 2015 to 2019 to explore more granular trends before the anomalies introduced by COVID-19.
-@sec-adapted shows the potential for this further investigation with a model consisting of the yearly data from 2015 to 2019 exhibiting a much stronger $R^2$ value of 0.875.
-This model is small (5 observations), thus a more extensive model involving the monthly data would be beneficial.
+@sec-results shows the potential for this further investigation with a model consisting of the yearly data from 2015 to 2019 exhibiting a much stronger $R^2$ value of 0.966 than the model in @sec-adapted.
+This model is small (5 observations), thus a more extensive model involving the monthly data would be necessary.
 Analyzing these pre-pandemic patterns could provide a clearer understanding of the role that metrics such as beds per 1,000, physician density, and attendance rates played during a relatively stable period.
@@
 
-## Full Analysis Dataset
+## Analysis Dataset
 
@@
 # Import the cleaned healthcare dataset
-cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+cleaned_data <- read_parquet("../data/02-analysis_data/healthcare_and_individual_treatments.parquet")
 
@@
 # Split and format the table into smaller parts
-part1 <- cleaned_data[, c(1:6)] %>% format_numeric()
-part2 <- cleaned_data[, c(6:12)] %>% format_numeric()
-part3 <- cleaned_data[, c(12:18)] %>% format_numeric()
+part1 <- cleaned_data[, c(1:5)] %>% format_numeric()
+part2 <- cleaned_data[, c(6:11)] %>% format_numeric()
+part3 <- cleaned_data[, c(12:17)] %>% format_numeric()
 part4 <- cleaned_data[, c(18:ncol(cleaned_data))] %>% format_numeric()
@@
 ### Original Model Residuals {#sec-residuals}
+
+```{r}
+#| echo: false
+#| warning: false
+#| eval: true
+#| label: fig-residadj
+#| tbl-cap: "Linear model explaining change in wait times based on healthcare system predictors (2015-2019)"
+# Extract residuals and fitted values from the model
+filtered_data <- filtered_data %>%
+  mutate(
+    residuals = residuals(modela),
+    fitted = fitted.values(modela)
+  )
+
+#### Diagnostic Plots ####
+
+# Plot b: Residuals vs Predictor 1 (attendance)
+plot_b <- ggplot(filtered_data, aes(x = attendance, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Attendance",
+    y = "Residuals",
+    x = "Attendance"
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Plot c: Residuals vs Predictor 2 (physicians_per_1000)
+plot_c <- ggplot(filtered_data, aes(x = physicians_per_1000, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Physicians per 1,000",
+    y = "Residuals",
+    x = "Physicians per 1,000"
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Plot d: Fitted values vs Actual values (Total)
+plot_e <- ggplot(filtered_data, aes(x = Total, y = fitted)) +
+  geom_point(color = "blue") +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
+  theme_classic() +
+  labs(
+    title = "Fitted vs Actual Values",
+    y = "Fitted Values",
+    x = "Actual Values"
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Plot e: Residuals vs Predictor 3 (beds_per_1000)
+plot_d <- ggplot(filtered_data, aes(x = beds_per_1000, y = residuals)) +
+  geom_point(color = "blue") +
+  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
+  theme_classic() +
+  labs(
+    title = "Residuals vs Beds per 1,000",
+    y = "Residuals",
+    x = "Beds per 1,000"
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Combine all plots into a grid
+combined_plot <- (plot_b | plot_c) / (plot_d | plot_e) +
+  plot_annotation(title = "Diagnostic Plots for Linear Model (2015-2019)")
+
+# Print the combined plot
+print(combined_plot)
+```
+@fig-residadj residuals show no clear patterns against attendance, physicians per 1,000, or beds per 1,000, supporting the assumption of linearity.
+There are no violations against constant variance, unlike in @fig-modeldetails.
+The fitted values reasonably align with the actual values.
+However, the limited data points (5 observations) reduce the robustness of these conclusions.
+
+\newpage
+
+### Adapted Model {#sec-adapted}
+```{r}
+#| echo: false
+#| warning: false
+#| eval: true
+#| label: tbl-modelresults
+#| tbl-cap: "Linear model explaining change in wait times based on healthcare system predictors"
+
+data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
+
+# Load the saved model
+model <- readRDS("../models/wait_times_model_2015_to_2022.rds")
+
+# Display the model summary in a table
+modelsummary::modelsummary(
+  list("Frequentist Linear Model" = model),
+  statistic = c("p.value")
+)
+```
+
+The linear regression model, presented in @tbl-modelresults, explores the relationships between healthcare system factors and the percentage change in wait times, though none of the variables demonstrate statistically significant effects, as indicated by p-values greater than 0.05.
+The number of hospital beds per 1,000 inhabitants shows a small negative effect ($\beta_1$=-0.556), suggesting a minimal decrease in wait times with higher bed availability, though this effect lacks statistical significance.
+Physician availability per 1,000 ($\beta_2$=2.517) indicates a larger positive association, yet it is also not statistically significant.
+Attendance rates ($\beta_3$ =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times.
+The model achieves an $R^2$ value of 0.599, indicating that approximately 60% of the variability in wait times is explained by the model, though the adjusted $R^2$ value of 0.299 reflects limited predictive power.
+With a sample size of n=8, the findings should be interpreted cautiously, as the lack of significance and low adjusted $R^2$ highlight potential limitations in the model's robustness and generalizability.
+
+\newpage
+
 ```{r}
@@
 
+# Add QQ Plot
+plot_qq <- ggplot(data, aes(sample = residuals)) +
+  stat_qq(color = "blue") +
+  stat_qq_line(color = "red", linetype = "dashed") +
+  theme_classic() +
+  labs(
+    title = "QQ Plot of Residuals",
+    y = "Sample Quantiles",
+    x = "Theoretical Quantiles"
+  ) +
+  theme(plot.title = element_text(size = 10))
+
 # Combine all plots into a grid
@@
 
-### Adapted Model {#sec-adapted}
-
-```{r}
-#| echo: false
-#| output: asis
-#| warning: false
-#| eval: true
-#| label: tbl-adj
-#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"
-
-#### Read data ####
-# Load the cleaned data
-analysis_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
-
-#### Filter data ####
-# Include only data from 2015 to 2019
-filtered_data <- analysis_data %>% 
-  filter(Year >= 2015 & Year <= 2019)
-
-#### Fit the linear model ####
-# Model: Percentage change in wait times as a function of healthcare predictors
-wait_times_model_adj <- lm(
-  Total ~ beds_per_1000 + physicians_per_1000 + attendance,
-  data = filtered_data
-)
-
-#### Save the model ####
-# Save the fitted model for reuse
-saveRDS(wait_times_model_adj, file = "../models/wait_times_model_pre_2020.rds")
-
-# Load the saved model
-modela <- readRDS("../models/wait_times_model_pre_2020.rds")
-
-# Display the model summary in a table
-modelsummary::modelsummary(
-  list("Frequentist Linear Model" = modela),
-  statistic = c("p.value")
-)
-```
-The linear regression model, summarized in @tbl-adj, examines the impact of healthcare system factors on wait times from 2015–2019.
-Hospital beds per 1,000 inhabitants ($\beta_1 =−176.99$) and physicians per 1,000 ($\beta_2 =−78.19$) both show negative associations with wait times, suggesting increased resources reduce delays, though neither effect is statistically significant ($p >0.05$).
-Attendance rates ($\beta_3=−0.66$) exhibit a minimal effect.
-The model explains 97.8% of the variability in wait times ($R^2$=0.978), though the adjusted $R^2$ of 0.911 reflects possible overfitting due to the small sample size ($n = 5$).
-While the model fits well, the lack of significant predictors highlights the need for further data to better capture drivers of wait times.
-The small RMSE of 0.58 indicates the linear trend is never far off from the actual wait times.
-
-\newpage
-
-```{r}
-#| echo: false
-#| warning: false
-#| eval: true
-#| label: fig-residadj
-#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"
-# Extract residuals and fitted values from the model
-filtered_data <- filtered_data %>%
-  mutate(
-    residuals = residuals(modela),
-    fitted = fitted.values(modela)
-  )
-
-#### Diagnostic Plots ####
-
-# Plot b: Residuals vs Predictor 1 (attendance)
-plot_b <- ggplot(filtered_data, aes(x = attendance, y = residuals)) +
-  geom_point(color = "blue") +
-  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
-  theme_classic() +
-  labs(
-    title = "Residuals vs Attendance",
-    y = "Residuals",
-    x = "Attendance"
-  ) +
-  theme(plot.title = element_text(size = 10))
-
-# Plot c: Residuals vs Predictor 2 (physicians_per_1000)
-plot_c <- ggplot(filtered_data, aes(x = physicians_per_1000, y = residuals)) +
-  geom_point(color = "blue") +
-  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
-  theme_classic() +
-  labs(
-    title = "Residuals vs Physicians per 1,000",
-    y = "Residuals",
-    x = "Physicians per 1,000"
-  ) +
-  theme(plot.title = element_text(size = 10))
-
-# Plot d: Fitted values vs Actual values (Total)
-plot_e <- ggplot(filtered_data, aes(x = Total, y = fitted)) +
-  geom_point(color = "blue") +
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
-  theme_classic() +
-  labs(
-    title = "Fitted vs Actual Values",
-    y = "Fitted Values",
-    x = "Actual Values"
-  ) +
-  theme(plot.title = element_text(size = 10))
-
-# Plot e: Residuals vs Predictor 3 (beds_per_1000)
-plot_d <- ggplot(filtered_data, aes(x = beds_per_1000, y = residuals)) +
-  geom_point(color = "blue") +
-  geom_hline(yintercept = 0, linetype = "dotted", color = "grey") +
-  theme_classic() +
-  labs(
-    title = "Residuals vs Beds per 1,000",
-    y = "Residuals",
-    x = "Beds per 1,000"
-  ) +
-  theme(plot.title = element_text(size = 10))
-
-# Combine all plots into a grid
-combined_plot <- (plot_b | plot_c) / (plot_d | plot_e) +
-  plot_annotation(title = "Diagnostic Plots for Linear Model (2015-2019)")
-
-# Print the combined plot
-print(combined_plot)
-```
-@fig-residadj residuals show no clear patterns against attendance, physicians per 1,000, or beds per 1,000, supporting the assumption of linearity.
-There are no violations against constant variance, unlike in @fig-modeldetails.
-The fitted values reasonably align with the actual values.
-However, the limited data points (5 observations) reduce the robustness of these conclusions.
-
-\newpage
-
 ```{r}
@@
 #| eval: true
-#| label: fig-adjmodelresults
-#| tbl-cap: "Projections of Linear Model onto Each Predictor (2015-2019)."
+#| label: fig-modelresults1
+#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining chane in wait times based on healthcare system predictors."
+
+# Load the saved model
+model <- readRDS("../models/wait_times_model_2015_to_2022.rds")
+
+# Load the dataset used for the model
+data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")
 
@@
       x = predictor_name,
-      y = "Wait Times (%)"
+      y = "Wait Times (#Weeks)"
     ) +
@@
 
-# Generate individual plots using filtered data (2015-2019)
-p1 <- create_projection_plot("attendance", filtered_data, modela)
-p2 <- create_projection_plot("physicians_per_1000", filtered_data, modela)
-p3 <- create_projection_plot("beds_per_1000", filtered_data, modela)
+# Generate individual plots
+p1 <- create_projection_plot("attendance", data, model)
+p2 <- create_projection_plot("physicians_per_1000", data, model)
+p3 <- create_projection_plot("beds_per_1000", data, model)
 
@@
   plot_annotation(
-    title = "Projections of Linear Model (2015-2019) onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
+    title = "Projections of Linear Model onto \nEach Predictor vs Average Wait Times Compared to 2015 ",
     caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
@@
 ```
-Overall, the linear model projections in @fig-adjmodelresults effectively captures general trends, particularly for hospital bed availability.
-However, the variability around the red lines, particularly for attendance and physicians per 1,000, indicates potential model limitations or missing interactions between predictors.
-The small sample size (n = 5) further limits the reliability of these insights, necessitating caution when generalizing these findings.
-Additional data or non-linear modeling may better address these discrepancies and provide a more comprehensive understanding of the relationships between these predictors and wait times.
-
-\newpage
+The projection plots in @fig-modelresults1 illustrate the linear model's predictions for the percentage change in wait times (`Total`) compared to the three predictors: `attendance`, `physicians_per_1000`, and `beds_per_1000`.
+The red lines depict the predicted relationships, while the blue points represent the observed data values.
+The `physicians_per_1000` predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend.
+Attendance also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance.
+In contrast, `beds_per_1000` shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend.
+For each plot of the predictor vs outcome variable there seems to be at least two data points which deviate significantly from the remaining points.
+These were hypothesized to be due to the outliers from trend witnessed in @fig-total stemming from the Covid-19 pandemic years.
+\newpage
+
+## Idealized Methodology
+
+### Limitations of Current NHS and OECD Data
+The NHS and OECD datasets utilized in this study provide valuable insights into hospital wait times through aggregate statistics such as median waiting times, hospital beds per capita, and physician density.
+These metrics are essential for maintaining patient confidentiality and offering a broad overview of healthcare system performance.
+However, these datasets exhibit limitations.
+They fail to capture detailed patient demographics, regional healthcare disparities, or systemic inefficiencies that may drive the increasing RTT (Referral to Treatment) wait times.
+Moreover, they omit subjective factors such as patient satisfaction, perceived urgency, and the emotional toll of delayed care—elements that could provide a more comprehensive understanding of the healthcare experience.
+
+Given this study's findings of no significant relationship between the predictors—hospital beds per 1,000, physicians per 1,000, and Type 1 Major Care Attendance Rates—and wait times, there is reason to explore more nuanced factors impacting delays.
+The absence of demographic data limits the ability to identify disparities in healthcare access and outcomes across different population groups.
+Furthermore, the lack of published records capturing patient perceptions of NHS wait times restricts the ability to evaluate how these delays are experienced by those affected.
+
+Addressing these gaps would require integrating more granular data on demographic variables, regional healthcare infrastructure, and patient-reported outcomes.
+Such data would enable a deeper analysis of the underlying drivers of wait times and support the development of targeted, effective interventions to improve healthcare delivery and patient experiences during the waiting period.
+
+### Overview
+To address the limitations of the current data, an idealized survey methodology is proposed to gather more granular and patient-centered data on hospital wait times and their systemic determinants.
+This methodology incorporates advanced sampling techniques, targeted recruitment strategies, and robust data validation measures to ensure a comprehensive and representative dataset.
+By focusing on both quantitative and qualitative aspects of healthcare delays, this approach aims to provide a deeper understanding of the factors influencing wait times and inform more effective policy interventions.
+
+### Sampling Method
+A stratified multi-stage sampling design will be employed to ensure representative data collection across the UK.
+This method captures the diversity of experiences in different regions, healthcare facilities, and patient populations.
+
+1. Stratification: The population will be divided into strata based on:
+
+    * Region: Urban, suburban, and rural areas.
+
+    * Demographics: Age, income, and ethnicity.
+
+    * Demographics: Age, income, and ethnicity.
+
+  For instance, one stratum might consist of elderly rural residents requiring orthopedic care, while another might focus on young urban adults accessing emergency services.
+
+2.
+Primary Sampling Units (PSUs): Regions within each stratum will be selected as PSUs using probability proportional to size (PPS).
+For example, a densely populated area like London will have a higher chance of selection compared to a sparsely populated rural region.
+
+3.
+Secondary Sampling Units (SSUs): Within each PSU, healthcare facilities (e.g., hospitals and clinics) will be chosen as SSUs.
+Facilities will be selected based on factors such as patient volume and type of services offered.
+
+4.
+Patient Selection: Patients within selected facilities will be randomly sampled, ensuring a balanced representation of various demographic and clinical profiles.
+Quotas will be used to ensure sufficient representation of underserved groups, such as low-income or minority patients.
+
+### Recruitment Strategy
+A multi-modal recruitment approach will maximize participation across diverse populations. Participants will be invited through:
+
+* **In-hospital Invitations**: Patients visiting selected facilities will be invited to participate, either during or after their appointments.
+
+* **Mail and Email**: Recruitment materials will be sent to patients identified through administrative records, with links to online surveys or options for telephone interviews.
+
+* **Community Outreach**: Partnerships with local organizations, especially in underserved areas, will help reach populations that might otherwise be excluded.
+
+* **Incentives**: Participants will receive small incentives, such as gift cards, to encourage participation and improve response rates.
+
+### Data Collection and Variables
+The survey will collect both quantitative and qualitative data:
+
+1. Demographic Information:
+
+    * Age, gender, income, education level, and ethnicity.
+
+    * Geographic location (postcode-level precision).
+
+2. Healthcare Access and Experiences:
+
+    * Type and urgency of medical procedures.
+
+    * Time elapsed between referral and treatment.
+
+    * Patient satisfaction with care received.
+
+3. Systemic and External Factors:
+
+    * Perceptions of healthcare resource adequacy (e.g., staffing, equipment).
+
+    * Preferences for care delivery models (e.g., in-person vs. virtual consultations).
+
+4. Perceived Impact of Delays:
+
+    * Emotional, physical, and financial effects of waiting for treatment.
+
+### Data Validation 
+The collected data will undergo validation to ensure accuracy and reliability:
+
+* Consistency Checks: Responses will be cross-validated with administrative records to ensure coherence (e.g., matching reported wait times with NHS data).
+
+* Non-Response Bias Analysis: Patterns of non-response will be analyzed, and statistical adjustments will be applied to account for underrepresented groups.
+
+* Pilot Testing: The survey will be tested on a small sample to refine questions, eliminate ambiguities, and improve response rates.
+
+### Simulation and Future Data Collection
+
+To complement the survey data, simulations will be conducted to model the impact of hypothetical interventions (e.g., increasing bed availability, optimizing physician allocation) on wait times.
+Additionally, future data collection efforts will incorporate continuous monitoring through partnerships with NHS facilities, enabling real-time updates and trend analysis.
+
+### Conclusion
+This idealized methodology addresses the critical gaps in existing data by integrating more nuanced information which cannot be collected by automated NHS portals, but rather surveying.
+The survey captures patient perspectives, systemic factors, and granular regional insights.
+By leveraging stratified sampling, a variety of recruitment strategies, and validation techniques, this approach can provide a more nuanced understanding of healthcare delays and patient perspectives in the UK.
+The findings will guide targeted interventions to improve healthcare access and efficiency, ensuring that resources are allocated where they are needed most.
+
+The survey is found at this link: <https://forms.gle/QJBeSUjxUejnSi3A9>.
+
+** *Copy of the Survey* **
+
+**NHS Patient Wait Time Experience Survey**
+
+This survey aims to gather insights into patients’ experiences with NHS referral-to-treatment (RTT) wait times, their perceptions of healthcare resources, and the impact of delays on their well-being.
+Your responses will help inform strategies to improve healthcare access and efficiency.
+Participation is voluntary, and all responses will remain confidential.
+
+Contact Information: For questions regarding this survey or the methodology used, please reach out to:
+
+-   **Chenika Bukes**, University of Toronto. Email: [chenika.bukes\@mail.utoronto.ca](malito:chenika.bukes@mail.utoronto.ca)
+
+**Section 1**: Demographics
+
+1. **What is your age?**
+
+  ```
+  - 18-25
+
+  - 25-34
+
+  - 35-44
+
+  - 45-54
+
+  - 55-64
+
+  - 65+
+  ```
+
+2. **What is your gender?**
+
+  ```
+  - Male
+
+  - Female
+
+  - Non-binary
+
+  - Prefer not to say
+
+  ```
+
+3. **What is your annual household income?**
+
+  ```
+  - Under £20,000
+
+  - £20,000-£39,999
+
+  - £40,000-£59,999
+
+  - £60,000-£79,999
+
+  - £80,000+
+
+  - Prefer not to say
+  ```
+
+4. **What is your ethnicity?**
+
+  ```
+  - White
+
+  - Black or Black British
+
+  - Asian or Asian British
+
+  - Mixed/Multiple ethnic groups
+
+  - Other
+
+  - Prefer not to say
+  ```
+
+1. **What is your post code?**
+
+  ```
+  Short answer text
+  ```
+
+**Section 2**: Healthcare Access
+
+5. **Have you had to wait longer than 18 weeks from referral for a medical procedure in the past 12 months?**
+
+  ```
+  - Yes
+
+  - No
+
+  ```
+
+5. **If yes, how long did you wait to be consulted after referral?**
+
+  ```
+  - 18-24 Weeks
+
+  - 24-30 Weeks
+
+  - 30-36 Weeks
+
+  - 36-42 Weeks
+
+  - 42-48 Weeks
+
+  - 48-52 Weeks
+
+  - 52+ Weeks
+
+  ```
+
+6. **What type of medical procedure did you wait for?**
+
+  ```
+  - General Surgery
+
+  - Orthopedics
+
+  - Cardiology
+
+  - ENT (Ear, Nose, Throat)
+
+  - Neurology
+
+  - Thoracic
+
+  - Rheumatology
+
+  - Plastic Surgery
+
+  - Geriatric
+
+  - Gynecology
+
+  - Opthamology
+
+  - Oral Surgery
+
+  - Cardiothoracic
+
+  - Neurosurgery
+
+  - Dermatology
+
+  - Gastroenterology
+  ```
+
+7. **If you waited for longer than 18 weeks for an appointment, were you contacted during the waiting period by an NHS representative for a check-up on your related health problem?**
+
+  ```
+  - Yes
+
+  - No
+  ```
+
+**Section 3:** Impact of Wait Times
+
+8. **Did the wait time affect your health or well-being?**
+
+  ```
+  - Yes, significantly
+
+  - Yes, moderately
+
+  - No
+  ```
+
+9. **If yes, how were you affected?**
+
+  ```
+  - Physical health deterioration
+
+  - Emotional distress or anxiety
+
+  - Financial burden
+
+  - Other...
+  ```
+
+10. **How do you feel about virtual consultations?**
+
+  ```
+  - Very satisfied
+
+  - Somewhat satisfied
+
+  - Impartial
+
+  - Unsatisfied
+
+  - Depends on the medical procedure in discussion
+
+  - Other...
+  ```
+
+**Section 4:** Perceptions of Healthcare System
+
+10. **How would you rate the adequacy of healthcare resources in your area?**
+  ```
+  - Excellent
+
+  - Good
+
+  - Average
+
+  - Poor
+
+  ```
+
+11. **What do you think would most improve healthcare wait times?**
+
+  ```
+  - More hospital beds
+
+  - More healthcare workers
+
+  - Better resource allocation
+
+  - Improved scheduling systems
+
+  - Policy changes
+  ```
+
+**Section 5**: Feedback and Suggestions
+
+12. **What changes would you like to see in the NHS to reduce referral to treatment waiting times?**
+
+13. **Do you have any suggestions to improve this survey?**
 

### 2024-12-03 scripts/02-clean_data.R

@@ -119,23 +119,10 @@ yearly_averages <- all_data %>%
 
 # Pivot the data so that years are rows and treatment functions are columns
 wide_yearly_averages <- yearly_averages %>%
-  pivot_wider(names_from = treatment_function, values_from = Average_Waiting_Time)
-
-# Calculate percent change from 2015 levels with robust handling of NA values
-wide_yearly_averages_percent_change <- wide_yearly_averages %>%
-  mutate(across(
-    -Year,  # Exclude the 'Year' column from the calculation
-    ~ ifelse(!is.na(.) & !is.na(first(.)), (.) / first(.) * 100 - 100, NA),  # Calculate percent change, handle NA
-    .names = "{.col}_percent_change"  # Rename columns to indicate percent change
-  ))
-
-# Drop the original columns and retain only the percent change columns
-wide_yearly_averages_percent_change <- wide_yearly_averages_percent_change %>%
-  select(Year, ends_with("_percent_change")) %>%
-  rename_with(~ gsub("_percent_change", "", .), ends_with("_percent_change"))  # Rename back to original names
+  pivot_wider(names_from = treatment_function, values_from = Average_Waiting_Time) 
 
 # Rename columns to keep only the first word for treatment names and resolve duplicates
-wide_yearly_averages_percent_change <- wide_yearly_averages_percent_change %>%
+wide_yearly_averages <- wide_yearly_averages %>%
   rename_with(~ {
     # Extract the first word
     renamed <- sapply(strsplit(., " "), `[`, 1)
@@ -203,7 +190,20 @@ a_and_e_data <- read.csv("./data/01-raw_data/a_and_e_activity_data_full.csv") |>
 final_data <- physician_data |>
   left_join(beds_data, by = "Year") |>
   left_join(a_and_e_data, by = "Year") |>
-  left_join(wide_yearly_averages_percent_change, by = "Year") |>
+  left_join(wide_yearly_averages, by = "Year") |>
+  select(
+    Year, 
+    beds_per_1000, 
+    physicians_per_1000, 
+    attendance,
+    Total
+  )
+
+
+final_data_full <- physician_data |>
+  left_join(beds_data, by = "Year") |>
+  left_join(a_and_e_data, by = "Year") |>
+  left_join(wide_yearly_averages, by = "Year") |>
   select(
     Year, 
     beds_per_1000, 
@@ -228,11 +228,10 @@ final_data <- physician_data |>
     Trauma,
[…]

### 2024-12-03 scripts/05-model_data.R

@@ -11,7 +11,7 @@ library(tidyverse)
 
 #### Read data ####
 # Load the cleaned data
-analysis_data <- read_csv("./data/02-analysis_data/final_healthcare_data.csv")
+analysis_data <- read_parquet("./data/02-analysis_data/final_healthcare_data.parquet")
 
 #### Fit the linear model ####
 # Model: Percentage change in wait times as a function of healthcare predictors
@@ -22,5 +22,21 @@ wait_times_model <- lm(
 
 #### Save the model ####
 # Save the fitted model for reuse
-saveRDS(wait_times_model, file = "./models/wait_times_model.rds")
+saveRDS(wait_times_model, file = "./models/wait_times_model_2015_to_2022.rds")
+
+#### Filter data for second model ####
+# Include only data from 2015 to 2019
+filtered_data <- analysis_data %>% 
+  filter(Year >= 2015 & Year <= 2019)
+
+#### Fit the linear model for 2015-2019####
+# Model: Percentage change in wait times as a function of healthcare predictors
+wait_times_model_adj <- lm(
+  Total ~ beds_per_1000 + physicians_per_1000 + attendance,
+  data = filtered_data
+)
+
+#### Save the model for 2015-2019####
+# Save the fitted model for reuse
+saveRDS(wait_times_model_adj, file = "./models/wait_times_model_pre_2020.rds")
 

### 2024-12-03 paper/paper.qmd
Updated data section in paper.
Finalized sub-section headings.

@@
 
-The NHS Referral to Treatment (RTT) waiting times dataset [@wait] is the primary source for waiting time statistics.
+The NHS Referral to Treatment (RTT) Waiting Times dataset [@wait] is the primary source for waiting time statistics.
 These data measure the median time (in weeks) between referral by a general practitioner (GP) and the start of treatment for elective care, capturing delays across 15 key medical procedures @tbl-treatments.
@@
 All data cleaning was done using `arrow` [@arrow], `tidyverse` [@tidyverse], `readxl` [@readxl], and `stringr` [@stringr].
-Below is a snapshot of the first five rows of the key data columns which will be used for this paper.
+Below is the cleaned data which will be used for this paper.
 @tbl-dataappendix provides the data including individual treatment area percentage change in waiting times from 2015 in table format.
@@
 This was followed by a sharp increase in 2021 and a steep surge in 2022, likely due to the backlog created during the pandemic, ongoing demand pressures, and resource constraints.
-These trends underscore the urgent need for targeted healthcare interventions, including expanded capacity, streamlined workflows, and policies to address the post-pandemic backlog and increasing demand.
 Overall, wait times have risen close to 30% since 2015, further emphasizing the need for targeted interventions to improve healthcare accessibility and efficiency.
@@
 However, a slight rebound is observed in 2021 and 2022, with the number of beds per 1,000 inhabitants stabilizing at just above 2.45.
-This recent stabilization could suggest efforts to address capacity constraints or temporary adjustments in response to the increased demand for hospital resources during the COVID-19 pandemic.
 The trend underscores the importance of evaluating the impact of long-term capacity changes on patient care and exploring strategies to ensure sufficient resources to meet population needs.
@@
 
-## Model set-up
+**Model Instantiation**
+
 The final model is displayed here:
@@
 
-### Model Assumptions {#sec-model-assumptions}
+**Model Assumptions**
+
 The linear regression model assumes:
@@
 
-### Model justification
+**Model justification**
 
@@
 
-### Hypotheses
+**Hypotheses**
+
 Each predictor variable is expected to contribute uniquely:
@@
 
-### Evaluation and Validation
+\newpage
+
+**Evaluation and Validation**
+
 To ensure the robustness of the findings, the model was validated through the following steps:
 
-1. Residual Diagnostics: We assessed residual plots for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
-2. Significance Testing: We conducted t-tests to assess whether each coefficient ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$) is statistically significant at the 0.05 level @sec-results.
+1. Residual Diagnostics: residual plots were assessed for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
+2. Significance Testing: t-tests were conducted to assess whether each coefficient ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$) is statistically significant at the 0.05 level @sec-results.
 3. Goodness of Fit: The R-squared statistic was calculated to measure how well the model explains the variability in wait times @sec-results.
@@
 Alternative Models: a variant containing data from 2015 to 2022 was tested in @sec-adapted to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit.
-This model had a much smaller $R^2$ and adjusted $R^2$ values and a much larger RMSE.
+This model had much smaller $R^2$ and adjusted $R^2$ values and a much larger RMSE.
 The observation size differences between the two models were not significant enough to account for the much poorer fit, thus the smaller model was kept for final analysis.
@@
 ```
-The projection plots in @fig-adjmodelresults illustrate the linear model's predictions for the average wait timein relation to the three predictors: attendance, physicians per 1,000, and beds per 1,000.
-The red lines in each plot represent the predicted relationships, while the blue points reflect the observed data.
-For attendance, the model suggests a negative relationship with wait times, as indicated by the downward-sloping red line.
-However, the observed data points display considerable variability around the prediction, with notable deviations suggesting the presence of unmodeled interactions or confounding factors.
-The relationship for physicians per 1,000 is also negative, with a steep downward slope showing that higher physician density is associated with reduced wait times.
-While the data points align more closely with the model's predictions compared to attendance, some deviations, particularly at lower physician densities, highlight potential limitations in the model’s assumptions.
-The strongest and most consistent relationship is observed for beds per 1,000, where the steep downward-sloping red line indicates a significant association between increased bed availability and reduced wait times.
-The observed data points align closely with this trend, suggesting that this predictor has a robust and straightforward influence on wait times.
-
-Across all predictors, however, the small sample size (n = 5) significantly limits the reliability and generalizability of the findings.
-While the data for 2015–2019 avoids the confounding impact of the COVID-19 pandemic, the limited dataset reduces the ability to draw definitive conclusions, and individual coefficient estimates remain non-significant despite the improved model fit.
-The variability around the predictions, particularly for attendance and physicians per 1,000, highlights potential missing interactions or non-linear relationships that the linear model cannot capture.
-The model overall captures the relationship between the predictors and the outcome variable better than the model investigated in @sec-adapted.
+The projection plots in @fig-adjmodelresults illustrate the linear model's predictions for average wait times in relation to the three predictors: attendance, physicians per 1,000, and beds per 1,000.
+Red lines represent the predicted relationships, while blue points reflect observed data.
+For attendance, the model suggests a negative relationship, but considerable variability and notable deviations hint at unmodeled interactions or confounding factors.
+For physicians per 1,000, the negative relationship is clearer, with a steeper downward slope, though deviations at lower densities reveal limitations in the model’s assumptions.
+The strongest relationship is seen for beds per 1,000, where a consistent downward slope indicates a significant association between increased bed availability and reduced wait times, with observed data aligning closely to this trend.
+
+However, the small sample size (n = 5) limits the reliability and generalizability of these findings.
+While the 2015–2019 data avoids the confounding effects of COVID-19, the limited dataset hampers definitive conclusions, and predictor coefficients remain non-significant.
+Variability in predictions for attendance and physicians per 1,000 suggests potential missing interactions or non-linear dynamics.
+Overall, the model better captures the relationship between predictors and wait times than the one analyzed in @sec-adapted.
 
@@
 
-### Limitations of Current NHS and OECD Data
+**Limitations of Current NHS and OECD Data**
+
 The NHS and OECD datasets utilized in this study provide valuable insights into hospital wait times through aggregate statistics such as median waiting times, hospital beds per capita, and physician density.
@@
 
-### Overview
+**Overview**
+
 To address the limitations of the current data, an idealized survey methodology is proposed to gather more granular and patient-centered data on hospital wait times and their systemic determinants.
@@
 
-### Sampling Method
+**Sampling Method**
+
 A stratified multi-stage sampling design will be employed to ensure representative data collection across the UK.
@@
 
-### Recruitment Strategy
+**Recruitment Strategy**
+
 A multi-modal recruitment approach will maximize participation across diverse populations. Participants will be invited through:
 
-* **In-hospital Invitations**: Patients visiting selected facilities will be invited to participate, either during or after their appointments.
-
-* **Mail and Email**: Recruitment materials will be sent to patients identified through administrative records, with links to online surveys or options for telephone interviews.
-
-* **Community Outreach**: Partnerships with local organizations, especially in underserved areas, will help reach populations that might otherwise be excluded.
-
-* **Incentives**: Participants will receive small incentives, such as gift cards, to encourage participation and improve response rates.
-
-### Data Collection and Variables
+* *In-hospital Invitations*: Patients visiting selected facilities will be invited to participate, either during or after their appointments.
+
+* *Mail and Email*: Recruitment materials will be sent to patients identified through administrative records, with links to online surveys or options for telephone interviews.
+
+* *Community Outreach*: Partnerships with local organizations, especially in underserved areas, will help reach populations that might otherwise be excluded.
+
+* *Incentives*: Participants will receive small incentives, such as gift cards, to encourage participation and improve response rates.
+
+**Data Collection and Variables**
+
 The survey will collect both quantitative and qualitative data:
@@
 
-### Data Validation 
+**Data Validation**
+
 The collected data will undergo validation to ensure accuracy and reliability:
@@
 
-### Simulation and Future Data Collection
+**Simulation and Future Data Collection**
 
@@
 
-### Conclusion
+**Conclusion**
+
 This idealized methodology addresses the critical gaps in existing data by integrating more nuanced information which cannot be collected by automated NHS portals, but rather surveying.
@@
 
-** *Copy of the Survey* **
-
-**NHS Patient Wait Time Experience Survey**
+**Copy of the Survey**
+
+*NHS Patient Wait Time Experience Survey:*
 

### 2024-12-03 paper/paper.qmd
Completed paper and README details.

@@
   labs(
-    title = "Median Number of Weeks for Total Wait Times Compared to 2015",
+    title = "Average Number of Weeks spent Waiting from Referral to Treatment",
     x = "Year",
-    y = "Average Wait Time (Median # of Weeks)",
+    y = "Average Wait Time (# of Weeks)",
     color = NULL
@@
 ### Data Cleaning {#sec-cleaning}
-The data cleaning process undertaken in this project involved several key steps to ensure the dataset was accurate, standardized, and ready for analysis.
-First, raw data on healthcare waiting times, physicians, hospital beds, and emergency admissions from various sources, such as the OECD and NHS, were loaded.
+The data cleaning process undertaken in this project involved several steps to ensure the dataset was accurate, standardized, and ready for analysis.
+First, raw data on healthcare waiting times, physicians, hospital beds, and emergency admissions from the OECD and NHS were loaded.
 For the waiting times data, all Excel files were processed by standardizing column names, extracting relevant fields (treatment function and average waiting time), and appending metadata such as the year and month derived from file names.
@@
 The data was then pivoted to make years the rows and treatment functions the columns, providing a wide-format structure suitable for further analysis.
-Special corrections were applied, such as manually imputing specific values for ENT in 2019 and 2020.
 
 All datasets were merged on the year variable to produce a consolidated dataset.
-Percent changes in waiting times from 2015 levels were computed for each treatment function to highlight trends over time.
+Percent changes in attendance rates from 2015 levels were computed to highlight trends over this study's time as opposed to the original percent changes from 2011 data.
 Lastly, columns with multiple-word treatment names were shortened to their first word, with duplicates (e.g., "General Surgery" and "General Medicine") resolved by appending unique suffixes.
@@
 
-    * Demographics: Age, income, and ethnicity.
-
   For instance, one stratum might consist of elderly rural residents requiring orthopedic care, while another might focus on young urban adults accessing emergency services.
@@
 
-2. **What is your gender?**
+2. **What sex were you assigned at birth, on your original birth certificate?**
 
@@
   - Female
-
-  - Non-binary
+  ```
+
+3. **How do you currently describe yourself (mark all that apply)?**
+
+  ```
+  - Female
+
+  - Male
+
+  - Transgender
+
+  - I use a different term [free-text]
+  ```
+
+4. **What is your annual household income?**
+
+  ```
+  - Under £20,000
+
+  - £20,000-£39,999
+
+  - £40,000-£59,999
+
+  - £60,000-£79,999
+
+  - £80,000+
 
   - Prefer not to say
-
-  ```
-
-3. **What is your annual household income?**
-
-  ```
-  - Under £20,000
-
-  - £20,000-£39,999
-
-  - £40,000-£59,999
-
-  - £60,000-£79,999
-
-  - £80,000+
+  ```
+
+5. **What is your ethnicity?**
+
+  ```
+  - White
+
+  - Black or Black British
+
+  - Asian or Asian British
+
+  - Mixed/Multiple ethnic groups
+
+  - Other
 
@@
 
-4. **What is your ethnicity?**
-
-  ```
-  - White
-
-  - Black or Black British
-
-  - Asian or Asian British
-
-  - Mixed/Multiple ethnic groups
-
-  - Other
-
-  - Prefer not to say
-  ```
-
-1. **What is your post code?**
+6. **What is your post code?**
 
@@
 
-5. **Have you had to wait longer than 18 weeks from referral for a medical procedure in the past 12 months?**
+7. **Have you had to wait longer than 18 weeks from referral for a medical procedure in the past 12 months?**
 
@@
 
-5. **If yes, how long did you wait to be consulted after referral?**
+8. **If yes, how long did you wait to be consulted after referral?**
 
@@
 
-6. **What type of medical procedure did you wait for?**
+9. **What type of medical procedure did you wait for?**
 
@@
 
-7. **If you waited for longer than 18 weeks for an appointment, were you contacted during the waiting period by an NHS representative for a check-up on your related health problem?**
+10. **If you waited for longer than 18 weeks for an appointment, were you contacted during the waiting period by an NHS representative for a check-up on your related health problem?**
 
@@
 
-8. **Did the wait time affect your health or well-being?**
+11. **Did the wait time affect your health or well-being?**
 
@@
 
-9. **If yes, how were you affected?**
+12. **If yes, how were you affected?**
 
@@
 
-10. **How do you feel about virtual consultations?**
+13. **How do you feel about virtual consultations?**
 
@@
 
-10. **How would you rate the adequacy of healthcare resources in your area?**
+14. **How would you rate the adequacy of healthcare resources in your area?**
   ```
@@
 
-11. **What do you think would most improve healthcare wait times?**
+15. **What do you think would most improve healthcare wait times?**
 
@@
 
-12. **What changes would you like to see in the NHS to reduce referral to treatment waiting times?**
-
-13. **Do you have any suggestions to improve this survey?**
+16. **What changes would you like to see in the NHS to reduce referral to treatment waiting times?**
+
+17. **Do you have any suggestions to improve this survey?**
+
+\newpage
 
